Establishing synthetic ribbon-type active zones in a heterologous expression system

Version 1

Abstract

Abstract

Encoding of several sensory modalities into neural signals is mediated by ribbon synapses. The synaptic ribbon tethers synaptic vesicles at the presynaptic active zone (AZ) and might act as a super-scaffold organizing AZ topography. Here we employed a synthetic biology approach to reconstitute ribbon-type AZs in HEK293 cells for probing their minimal molecular requirements and studying presynaptic Ca2+ channel clustering. Co-expressing a membrane-targeted version of the AZ-protein Bassoon and the ribbon core protein RIBEYE, we observed structures recapitulating basic aspects of ribbon-type AZs, which we call synthetic ribbons or SyRibbons. SyRibbons with Ca2+ channel clusters formed upon additional expression of CaV1.3 Ca2+ channels and RIM-binding protein 2 (RBP2), known to promote presynaptic Ca2+ channel clustering. Confocal and super-resolution microscopy along with functional analysis by patch-clamp and Ca2+-imaging revealed striking similarities and interesting differences of SyRibbons in comparison to native IHC ribbon-type AZs. In summary, we identify Ca2+ channels, RBP, membrane-anchored Bassoon, and RIBEYE as minimal components for reconstituting a basic ribbon-type AZ. SyRibbons might complement animal studies on molecular interactions of AZ proteins.

Introduction

Synapses of sensory receptor cells of the inner ear and the retina are hallmarked by the presence of an electron-dense structure called the synaptic ribbon, which tethers synaptic vesicles (SVs) at the presynaptic active zone (AZ, [Moser et al, 2019; Matthews & Fuchs, 2010]). Proposed functional roles of the ribbon include (i) its role as a SV replenishment machine for tireless neurotransmission (Bunt, 1971; Holt et al, 2004; Khimich et al, 2005; Frank et al, 2010; Snellman et al, 2011; Graydon et al, 2011; Vaithianathan et al, 2016), (ii) “super-scaffold” regulating the abundance, topography, and function of AZ players such Ca2+ channels and release sites and their tight coupling (Khimich et al, 2005; Frank et al, 2010; Wong et al, 2014; Maxeiner et al, 2016; Jean et al, 2018; Grabner & Moser, 2021), and iii) coordination of SV release (Heidelberger et al, 1994; Glowatzki & Fuchs, 2002; Singer et al, 2004; Edmonds, 2004; Mehta et al, 2013). Yet, our molecular understanding of how these postulated functions are executed by synaptic ribbons is still limited.

Likewise, the clustering of Ca2+ channels at the AZ, how it is varied among different or even the same synapse types and how the ribbon contributes towards this, remain active areas of research. For example, ribbon synapses of mature inner hair cells (IHC) assemble CaV1.3 voltage-gated Ca2+ channels at the base of the ribbon, but vary in shape, size and channel complement (20-330) of the CaV1.3 clusters (Frank et al, 2010; Wong et al, 2014; Neef et al, 2018) partially dependent on the position within the IHC (Ohn et al, 2016; Özçete & Moser, 2021). IHC synapses lacking AZ-anchored ribbons in Bassoon mutant mice showed fewer Ca2+ channels with altered topography (Frank et al, 2010; Neef et al, 2018). While genetic disruption of Ca2+ channel-tethering AZ proteins RIM2 (Jung et al, 2015) and RBP2 (Krinner et al, 2017) reduced the abundance of Ca2+ channels at IHC synapses, their topography retained the typical stripe-like shape of Ca2+ clusters. Constitutive deletion of RIBEYE, the core scaffold protein of the synaptic ribbon, provided, so far, the most direct test of the scaffolding role of the ribbon (Maxeiner et al, 2016; Jean et al, 2018; Becker et al, 2018; Grabner & Moser, 2021). Here, ribbon-less synapses of IHCs showed an altered Ca2+ channel topography with a more spatially widespread presynaptic Ca2+ signal but intact Ca2+ channel complement (Jean et al, 2018). Moreover, the voltage-dependence of activation as well as the inactivation of Ca2+ channels appeared to be altered, suggesting a possible role of the synaptic ribbon in also regulating Ca2+ channel physiology (Jean et al, 2018). Changes in Ca2+ channel physiology upon deletion of RIBEYE were also found for CaV1.4 at ribbon-less rod photoreceptor AZs (Grabner & Moser, 2021). Finally, IHCs lacking piccolino, a ribbon-synapse specific isoform of the AZ protein piccolo, show AZs with smaller synaptic ribbons and altered clustering yet normal complement of Ca2+ channels (Michanski et al, 2023).

So how does the synaptic ribbon tune into the orchestra of AZ proteins that collectively organize Ca2+ channel abundance, topography, and function? To what extent can the effects observed upon disruption of Bassoon be attributed to the concomitant loss of the ribbon? These questions seem even more relevant given that, so far to our knowledge, no direct interactions between Bassoon (Frank et al, 2010) or RIBEYE and the Ca2+ channel have been described. Here, we adopted a bottom-up synthetic biology approach to assemble a minimal presynaptic protein machinery required to bring together RIBEYE and Ca2+ channels in a heterologous expression system to assess the role of the synaptic ribbon in regulating Ca2+ channel clustering and physiology. Similar approaches have previously been employed for reconstituting aspects of conventional synapses, for instance in “hemisynapses” between co-cultured neurons and non-neuronal cells overexpressing presynaptic or postsynaptic components (for review see Craig et al, 2006). More recently, Munc13-1 supra-molecular assemblies which recruit release machinery proteins were reconstituted (Sakamoto et al, 2018).

Expression of RIBEYE in synapse-naïve cell lines such as COS-7 cells led to large cytosolic assemblies (Schmitz et al, 2000). Membrane-proximal RIBEYE clusters and ribbon-like electron dense structures with vesicles could be found in a retinal progenitor cell-line (R28, [Magupalli et al, 2008]) yet, these cells might not be considered synapse-naïve. Regardless, the presence and function of Ca2+ channels at synthetic ribbon synapses have not yet been studied. The need for co-expressing at least three Ca2+ channel subunits (CaVαx, CaVβx and CaVα2δx) in synapse-naïve cell lines renders acute co-expression of AZ multidomain proteins cumbersome. Here we took advantage of synapse-naïve Human Embryonic Kidney 293 (HEK293) cells stably expressing inducible CaV1.3α1, which is the predominant subtype of voltage-gated Ca2+ channels at IHC ribbon synapses (Brandt et al, 2003; Platzer et al, 2000; Dou et al, 2004), as well as constitutively expressed CaVβ3 and CaVα2δ1. We co-expressed RIBEYE along with membrane targeted Bassoon and observed structures with striking resemblance to IHC synaptic ribbon-type AZs. We characterized the structure and function of these synthetic ribbon-type AZs and identified Bassoon, RBP2, RIBEYE and CaV1.3 channels as the minimal components required for assembling a basic ribbon-type AZ. We observed that synthetic ribbons recruit CaV1.3 channels in large clusters and show localized Ca2+ influx upon stimulation. Our results support the role of synaptic ribbons in promoting CaV1.3 channel clustering. Synthetic ribbon-type AZs offer a novel approach for functionally studying protein-protein interaction and will likely complement, refine and reduce experiments on native ribbon synapses.

Results

Membrane-targeted Bassoon recruits RIBEYE to the cell membrane in HEK293 cells where structures alike inner hair cell synaptic ribbons are formed

Human epithelial HEK293 cells have the advantage of not expressing the synaptic machinery components studied here. This provides a clean background for reconstituting synthetic synapses from a minimal set of synaptic proteins. Our first step towards assembling a ribbon-type AZ in a heterologous expression system was to express RIBEYE, the core scaffold protein of the synaptic ribbon, and target it to the cell membrane. We performed transient transfection of a RIBEYE construct with a C-terminal EGFP tag in HEK293 cells and observed large spherical clusters of RIBEYE that appeared largely cytosolic (Fig 1Aii), in contrast to a diffuse cytosolic distribution when merely expressing EGFP (Fig 1Ai). These RIBEYE clusters form due to self-assembling properties of RIBEYE via multiple sites of homophilic interaction as have been demonstrated before in several cell lines. They do not colocalize with the endoplasmic reticulum (ER), Golgi apparatus, or lysosomes and hence, are unlikely to reflect RIBEYE entrapped in protein trafficking pathways or to represent degradation products of overexpressed RIBEYE (Fig EV1 A, B and C).

Membrane targeting of Bassoon using a palmitoylation consensus sequence.

A Representative confocal images of HEK293 cells transfected with (i) GFP only (CtBP2/RIBEYE in magenta labeling nuclei, GFP in cyan), (ii) RIBEYE-GFP (CtBP2/RIBEYE in magenta, GFP in cyan), (iii) Bassoon (green) and (iv) RIBEYE-GFP and Bassoon. Note that antibodies against RIBEYE B-domain and the nuclear transcription factor CtBP2 result in similar staining patterns as CtBP2 is transcribed from the same gene as RIBEYE and is identical to the RIBEYE-B domain except for the first 20 N-terminal amino acids. Upper panel shows maximum projections and lower panel shows exemplary single sections from confocal stacks. Scale bar = 5 µm.

B Schematic of construct for membrane targeting of Bassoon. The first 95 amino acids from full-length Bassoon were replaced with a palmitoylation consensus sequence from GAP43. Constructs without and with a C-terminal GFP tag were used, as depicted.

C Sample confocal image (single section) showing membrane targeted palm-Bassoon (green) expressed in HEK293 cells appears as puncta distributed along the periphery of the cell, marked by Na, K-ATPase α1(blue). Inset shows maximal projection of confocal section. Scale bar = 5µm.

D Zoom-in from (C) shows colocalisation of palm-Bassoon with membrane marker Na, K-ATPase α1. Scale bar = 1µm.

E Quantification of Bassoon signal intensity at periphery vs inside of cell. Cells expressing palm-Bassoon (N = 10 cells) clearly show a higher peripheral distribution compared to cells expressing full-length Bassoon (N = 10 cells), ****P < 0.0001, Mann-Whitney-Wilcoxon test. Overlaid data points represent individual cells, crosses represent the mean values, central band indicates the median, whiskers represent 90/10 percentiles, and boxes represent 75/25 percentiles.

Subcellular distribution of RIBEYE clusters in HEK293 cells.

Maximal projections of confocal sections of HEK293 cells transfected with RIBEYE-GFP (green). Representative co-staining for (A) calnexin (endoplasmic reticulum marker), (B) GM130 (cis-Golgi marker) and (C) LAMP1 (lysosomal marker) have been shown respectively (blue). RIBEYE clusters do not appear to localize in either of the three compartments. Scale bar = 10 µm.

Next, for membrane targeting of these cytosolic RIBEYE clusters, we co-expressed Bassoon (Fig 1Aiii). Prior work on the molecular underpinnings of ribbon synapses had identified the multidomain cytomatrix of the active zone protein Bassoon (tom Dieck et al, 1998) to critically contribute to anchoring the synaptic ribbon to the AZ membrane (Khimich et al, 2005; Dick et al, 2003; tom Dieck et al, 2005). Co-expression of full-length Bassoon along with RIBEYE in HEK293 cells showed colocalizing clusters of the two proteins that, however, remained largely cytosolic (Fig 1Aiv).

Next, for plasma membrane-targeting of Bassoon, we generated a construct by removing the first 95 N-terminal amino acids of Bassoon and replacing these with a palmitoylation consensus sequence of the neuronal Growth Associated Protein 43 (GAP43). We refer to this as palm-Bassoon throughout, and we used constructs with and without a C-terminal EGFP tag (Fig 1B). Expression of either of these palm-Bassoon constructs in HEK293 cells showed comparable immunofluorescence patterns with Bassoon puncta spread across the periphery of the cell, largely colocalizing with the endogenously expressed membrane protein Na, K-ATPase α1 (data representative of 3 transfections, Fig 1C, D). Comparing the ratio of Bassoon signal intensity at the periphery versus inside of the cell in randomly selected single sections from confocal stacks of palm-Bassoon and Bassoon transfected cells (N = 10 cells, 3 transfections per group, Fig 1E) demonstrated a higher Bassoon signal intensity at the periphery of cells expressing the palm-Bassoon construct (****P < 0.0001, Mann-Whitney-Wilcoxon test), implying successful plasma membrane targeting of Bassoon.

Next, we co-expressed palm-Bassoon and RIBEYE in HEK293 cells and observed colocalizing RIBEYE and Bassoon immunofluorescent puncta at the periphery of cells (7 transfections; Fig 2Ai). Closer inspection of these immunofluorescent puncta with STimulated Emission Depletion (STED) nanoscopy (Fig 2Aii) revealed discrete structures typically consisting of ellipsoidal RIBEYE clusters juxtaposing on top of plate-like palm-Bassoon structures which seemingly anchor the RIBEYE clusters to the plasma membrane. Interestingly, we found the morphology of the RIBEYE + palm-Bassoon structures to be strikingly reminiscent of the arrangement of the two proteins in IHC ribbon synapses, where an ellipsoid/spherical synaptic ribbon composed of RIBEYE is found seated on a Bassoon plate that anchors it to the presynaptic AZ membrane (e.g. Wong et al, 2014; Michanski et al, 2019, 2023) (Fig 2B, i and ii; data from Michanski et al, 2023).

Co-expression of RIBEYE with palm-Bassoon results in ribbon-type AZ-like structures

A (i) Representative confocal image (single section) of a HEK293 cell transfected with RIBEYE-GFP (magenta) and palm-Bassoon (green). Co-expression of RIBEYE with palm-Bassoon targets RIBEYE to the cell membrane. Inset shows maximum projection. Scale bar = 5µm. (ii) Exemplary 2-D STED images for RIBEYE – palm-Bassoon juxtapositions acquired from cells as shown in (i); Scale bar = 500nm; individual channels have been depicted with an intensity-coded look-up table with warmer colors indicating higher intensity.

B (i) Representative maximum projection of confocal sections of apical organ of Corti from a Wistar rat (postnatal day 18); data as published in Michanski et al, 2023, stainings were for CtBP2/RIBEYE (magenta; labeling synaptic ribbons and IHC nuclei) and Bassoon (green; spots juxtaposing with ribbons represent IHC AZs, spots not juxtaposing with ribbons represent efferent synapses formed by lateral olivocochlear neurons onto SGN boutons). Scale bar = 5µm. (ii) Juxtaposing RIBEYE and Bassoon spots imaged in 2-D STED and confocal mode respectively. Scale bar = 500nm. Note the striking resemblance to reconstituted RIBEYE + palm-Bassoon structures in HEK293 cells as shown in (Aii). STED images are from 3 sample transfections, representative of 7 total transfections; individual channels have been depicted with an intensity-coded look-up table with warmer colors indicating higher intensity.

C Quantification of RIBEYE and Bassoon signal intensity at periphery vs inside of cell shows a higher peripheral distribution of RIBEYE and Bassoon in HEK293 + RIBEYE + palm-Bassoon cells (N=9 cells) as compared to HEK293 + RIBEYE + Bassoon cells (N=9 cells); ****P < 0.0001, Mann-Whitney-Wilcoxon test. Overlaid data points represent individual cells, crosses represent mean values, central band indicates the median, whiskers represent 90/10 percentiles, and boxes represent 75/25 percentiles.

D Distribution of volumes of RIBEYE and palm-Bassoon puncta from HEK293 cells expressing RIBEYE and palm-Bassoon (n = 20 cells, quantifications from 5 sample transfections). Volumes of synaptic ribbons from rat IHCs (n = 29 cells, 3 rats) have been plotted for comparison.

E Box plot depicting data from (D). Volumes of RIBEYE and palm-Bassoon puncta are comparable to each other (P > 0.99, Kruskal-Wallis test with post hoc Dunn’s multi-comparison), but on an average much smaller and considerably more variable when compared to volumes of RIBEYE and Bassoon puncta from rat IHCs respectively (****P < 0.0001, Kruskal-Wallis test with post hoc Dunn’s multi-comparison). Overlaid plus signs represent individual spots, crosses represent mean values, central band indicates the median, whiskers represent 90/10 percentiles and boxes represent 75/25 percentiles.

F RIBEYE puncta in HEK cells expressing RIBEYE and palm-Bassoon (nspots = 864, N = 20 cells) appear more spherical than Bassoon puncta in the same cells (nspots = 961, N = 20 cells; ****P < 0.0001, Kruskal-Wallis test with post hoc Dunn’s multiple comparison test). Note the similar trend in IHC synaptic ribbons where RIBEYE puncta are more spherical than Bassoon puncta (nspots = 290 for RIBEYE and nspots = 314 for Bassoon, N = 29 cells, 3 rats; ****P < 0.0001, Kruskal-Wallis test with post hoc Dunn’s multiple comparison test). Overlaid plus signs represent individual spots, crosses represent mean values, central band indicates the median, whiskers represent 90/10 percentiles and boxes represent 75/25 percentiles.

We compared the RIBEYE and Bassoon signal intensities at the periphery versus inside of the cell in randomly selected single sections from confocal stacks of HEK cells expressing RIBEYE and palm-Bassoon (N = 9 cells) and HEK cells expressing RIBEYE and full-length Bassoon (N=9 cells). We found an increased peripheral distribution of both RIBEYE and Bassoon when using palm-Bassoon (Fig 2C, ****P < 0.0001, Mann-Whitney-Wilcoxon test), implying successful membrane targeting of RIBEYE by palm-Bassoon. In each transfection, ∼10% cells showed co-expression of both RIBEYE and palm-Bassoon (for representative sample overview see Fig EV2). Of those, cells expressing discrete RIBEYE-palm Bassoon structures were discernible by the characteristic RIBEYE distribution along the periphery of the cell. This peripheral distribution in turn seemingly depends upon RIBEYE and palm-Bassoon expression ratios (shown in Figure EV2B) as cells with little to no palm-Bassoon expression show predominantly cytosolic RIBEYE puncta and were not used for analysis. For simplicity, we henceforth refer to the structures composed of RIBEYE and palm-Bassoon in HEK293 cells as SyRibbons (for synthetic ribbons). We next performed three-dimensional surface renderings using Imaris 9.6 (Oxford Instruments) to assess these structures. In a given cell, only structures with colocalizing RIBEYE and palm-Bassoon immunofluorescence were considered for analysis to exclude the occasional non-membrane localized spots (on average 51.75 ± 40.84 RIBEYE surfaces colocalizing with Bassoon surfaces per cell; N = 20 cells). The volumes of RIBEYE and palm-Bassoon surfaces of SyRibbons were smaller on average and more variable (average volume ± standard deviation (SD) = 0.19 ± 0.23 µm3 with a coefficient of variation (CV) = 1.23 for RIBEYE; nspots = 864 and volume = 0.17 ± 0.18 µm3 with CV = 1.10 for Bassoon; nspots = 951; data from N = 20 cells, quantifications from 5 sample transfections) than volumes of synaptic ribbons and Bassoon immunofluorescent puncta from rat IHCs (volume = 0.29 ± 0.17 µm3, CV = 0.59 for RIBEYE; nspots = 290, and volume = 0.27 ± 0.18 µm3, CV = 0.67 for Bassoon; nspots = 314, data from N = 29 cells), (Figure 2D, E). Moreover, volumes of RIBEYE surfaces show a high positive correlation to volumes of corresponding palm-Bassoon surfaces (Pr = 0.778, ****P < 0.0001), implying larger palm-Bassoon structures may recruit bigger RIBEYE structures to the plasma membrane. We also note that volume of RIBEYE surfaces in SyRibbons appears smaller and well-regulated in contrast to the predominantly large, cytosolic RIBEYE assemblies in cells co-expressing RIBEYE and full-length Bassoon (volume = 0.30 ± 0.36 µm3, nspots = 460, N = 10 cells, ****P < 0.001, Mann-Whitney-Wilcoxon Test). In turn, Bassoon clusters seemed regulated by co-expressed RIBEYE: Bassoon surfaces at the plasma membrane were larger at SyRibbons than in the absence of RIBEYE in cells only expressing palm-Bassoon (volume = 0.14 ± 0.21 µm3, nspots = 2217, N = 13 cells, ****P < 0.001, Mann-Whitney-Wilcoxon Test).

Nonetheless, the fact that next to structures with volumes comparable to IHC ribbons, we also encountered smaller and larger structures, likely reflects poorer regulation of RIBEYE and palm-Bassoon expression in the heterologous system. RIBEYE clusters constituting SyRibbons appeared more spherical than the plate-like Bassoon structures in the same cells (sphericity Ψ = 0.91 ± 0.05, nspots = 864 for RIBEYE versus Ψ = 0.86 ± 0.07, nspots= 951 for Bassoon, data from N = 20 cells; ****P < 0.0001, Kruskal-Wallis test with post hoc Dunn’s multiple comparison test). This follows the same trend as in rat IHCs where RIBEYE spots indeed appear more spherical (Ψ = 0.90 ± 0.05, nspots = 290) compared to Bassoon spots (Ψ = 0.76 ± 0.10, nspots = 314); ****P < 0.0001, Kruskal-Wallis test with post hoc Dunn’s multiple comparison test, data from N = 29 cells (Figure 2F).

Membrane localization of RIBEYE depends on expression levels of palm-Bassoon

A Exemplary overview image of HEK293 cells co-transfected with RIBEYE (magenta) and palm-Bassoon (green) showing how to identify cells with SyRibbons. On an average about 10% of cells appear co-transfected of which some show peripheral distribution of RIBEYE (examples marked by arrows). Scale bar = 100 µm for overview and 20 µm for zoom-in.

B Peripheral distribution of RIBEYE appears to be dependent on expression levels of palm-Bassoon. Note the cell in the top panel with little to no palm-Bassoon expression and predominantly cytosolic RIBEYE puncta. Scale bar = 10 µm.

SyRibbons employ RBP2 to tether CaV1.3 channels.

To study the potential of SyRibbons to cluster Ca2+ channels, we employed HEK293 cells stably expressing an inducible transgene of CaV1.3α1 along with constitutive transgenes for CaVβ3 and CaVα2δ1. We either used tetracycline for inducible CaV1.3α1 channel expression (Fig 3A) or performed transient transfections with CaV1.3 constructs containing an N-terminal EGFP- (Fig 3B) or Halo-tag (Fig 3C) for direct fluorescence imaging. Expression of either Ca2+ channel complex resulted in small clusters at the plasma membrane as represented in Fig 3D (average volume 0.08 ± 0.13 µm3, nspots = 949, N = 12 cells for untagged inducible CaV1.3α1; 0.07 ± 0.08 µm3, nspots = 392, N = 11 cells for EGFP-CaV1.3; and 0.07 ± 0.09 µm3, nspots = 1281, N = 16 cells for Halo-CaV1.3α; Puntagged/EGFP = 0.547, Puntagged/Halo > 0.999, PHalo/EGFP = 0.582, Kruskal-Wallis test with post hoc Dunn’s multiple comparison test).

HEK293 cells expressing CaV1.3

A Representative confocal section of a HEK293 cell stably expressing an inducible CaV1.3α transgene (untagged) along with constitutive transgenes for CaVβ3 and CaVα2δ1. Immunostainings have been performed using antibodies against CaV1.3 (green) and Na, K-ATPase α1 (blue, labeling the plasma membrane). Scale bar = 10 µm.

B Representative confocal section of a HEK293 cell stably expressing CaVβ3 and CaVα2δ1, and transiently transfected with an N-terminal EGFP-tagged CaV1.3 construct. Immunostainings have been performed using antibodies against GFP (green) and Na, K-ATPase α1 (blue). Scale bar = 10 µm.

C Representative confocal section of a HEK293 cell stably expressing CaVβ3 and CaVα2δ1 and transiently transfected with an N-terminal Halo-tagged CaV1.3 construct. Immunostainings have been performed using antibodies against CaV1.3 (green) and Halo tag (magenta). Scale bar = 10 µm.

D Expression of either of the Ca2+ channel complexes result in clusters of comparable sizes, (P > 0.05, Kruskal-Wallis test with post hoc Dunn’s multiple comparison test). Overlaid plus signs represent individual spots, crosses represent mean values, central band indicates the median, whiskers represent 90/10 percentiles and boxes represent 75/25 percentiles.

Co-expression of palm-Bassoon did not show colocalization with CaV1.3, which is in line with observations with full-length Bassoon (Frank et al, 2010) and the same applied when co-expressing RIBEYE and CaV1.3, arguing against a direct interaction for both (Figure EV3). Rab-interacting molecule-binding protein (RIM-BP or RBP) links Bassoon to Ca2+ channels (Davydova et al, 2014) and hence was an interesting candidate for tethering CaV1.3 to Bassoon clusters. RBP has been shown to interact with CaV1.3 (Hibino et al, 2002) and to be required for normal CaV1.3 Ca2+ channel clustering at the IHC ribbon synapse (Krinner et al, 2017, 2021). Indeed, when co-expressing RBP2 with palm-Bassoon and CaV1.3, we found juxtaposed palm-Bassoon and CaV1.3 both with immunolabeled (Fig. 4A) and live-labeled (Fig. 4B, C) proteins indicating successful clustering of CaV1.3 at synthetic AZs.

CaV1.3 does not appear to colocalize directly with RIBEYE or with palm-Bassoon

A Representative confocal section of a HEK293 cell transfected with Halo-CaV1.3 (blue) and palm-Bassoon (green). The two proteins do not appear to colocalize. Scale bar = 5 µm.

B Representative confocal section of a HEK293 cell transfected with Halo-CaV1.3 (blue) and RIBEYE-GFP (red). RIBEYE and CaV1.3 also do not appear to colocalize, speaking against a possible direct interaction between the two proteins. Scale bar = 5 µm.

RBP2 bridges palm-Bassoon and CaV1.3 to form supra-molecular assemblies at the plasma membrane.

A Representative confocal images (single sections) of HEK293 cells (fixed) transfected with palm-Bassoon (green), RBP2 (red) and N-terminal Halo-tagged CaV1.3 (blue). The three colocalizing proteins appear to form supra-molecular assemblies at the cell membrane. Scale bar = 5 µm for left panel showing cell overviews, 2 µm for right panels showing zoom-ins.

B Representative confocal images (single sections) of live HEK293 cells transfected with palm-Bassoon-GFP (green), RBP2-p2A-mKATE2 (red, representing only mKATE2 signal and not RBP2 localization) and Halo-CaV1.3 (blue, labeled by with JF-646 HaloTag Ligand before imaging). Scale bar = 5 µm.

C Exemplary super-resolution STED images from live-labeled samples in (B) showing juxtaposition of palm-Bassoon and Halo-CaV1.3 signal at the plasma membrane in RBP2 positive cells. Scale bar = 1 µm.

We next performed ruptured whole-cell patch-clamp to assess if RBP2 and/or palm-Bassoon or RIBEYE co-expression results in changes in CaV1.3 Ca2+ currents. After tetracycline treatment for CaV1.3α1 expression (18-24 hours), cells were transfected with both RBP2-p2A-mKATE2 and palm-Bassoon-GFP, with only palm-Bassoon-GFP, only RBP2-p2A-mKATE2 or RIBEYE-GFP constructs (Figure 5A). After 18-24 hours, the cell culture media was changed, and cells were recorded 24-36 hours afterwards. We used 10 mM [Ca2+]e and recorded Ca2+ current (density) - voltage relations (IVs) ∼1 minute after establishing the whole-cell configuration by applying step depolarizations of 20 ms from −86.2 mV to 58.8 mV in 5 mV increments. Ca2+ current density was on an average approximately twice as large in cells expressing both RBP2 and palm-Bassoon when compared with induced-only controls (Figure 5B, D; **P = 0.0041, Kruskal-Wallis test with post hoc Dunn’s multiple comparison test). Ca2+ current densities also tended to be larger for cells expressing only palm-Bassoon or RBP2, but this difference was not statistically significant (P = 0.374 and 0.164 respectively, Kruskal-Wallis test with post hoc Dunn’s multiple comparison test), while for cells expressing RIBEYE there was no trend to differ (P > 0.999, Kruskal-Wallis test with post hoc Dunn’s multiple comparison test). We did not observe any noticeable differences in voltage-dependence, activation kinetics or inactivation kinetics of CaV1.3 Ca2+ current upon expression of any of these AZ proteins (Figure 5C, E, F; Figure EV4A, B).

Whole-cell Ca2+ current amplitudes increase upon co-expression of RBP2 and palm-Bassoon.

A Impact of co-expression of different ribbon-type AZ proteins on CaV1.3α1/β3/α2δ1 Ca2+ currents. Color code: non-transfected controls expressing only CaV1.3 (black), co-expression of RBP2 + palm-Bassoon (data depicted in turquoise), only palm-Bassoon (green), only RBP2 (blue) or only RIBEYE (magenta). Numbers of recorded cells/group have been indicated in the figure (bottom right).

B Co-expression of RBP2 and palm-Bassoon seemingly increases the whole-cell Ca2+ current density in HEK293 cells expressing CaV1.3, [Ca]e = 10mM. Expression of palm-Bassoon or RBP2 individually also results in a mild increase, but this trend is not statistically significant as shown in (D). Expression of RIBEYE by itself causes no change in Ca2+ current density. Plots show lines representing mean Ca2+ current density traces and shaded area represents ± SEM.

C We fitted a Boltzmann function to traces in (B) to determine the fractional activation of CaV1.3 channels. The voltage-dependence of Ca2+ current influx does not appear to be different for any of the four scenarios.

D Box plot depicting peak Ca2+ current density values from (A). P values are as follows: **PCaV only/CaV+RBP2+PBsn = 0.0041, PCaV only/CaV+PBsn = 0.3737, PCaV only/CaV+RBP2 = 0.1643, PCaV only/CaV+RIBEYE > 0.999; Kruskal-Wallis test with post hoc Dunn’s multiple comparison test.

E,F Box plots depicting voltage of half-maximal activation (Vhalf) and slope (k) of the Boltzmann fits in (C) show no statistically significant differences.

Kinetics of CaV1.3 channels are not altered upon co-expression of various AZ proteins.

A Average peak amplitude normalized Ca2+ current traces in response to depolarizations of 500 ms from −91.2 to −21.2 mV (shaded area represents ± SEM). Expression of palm-Bassoon+RBP2 (turquoise), only palm-Bassoon (green), RBP2 (blue) or RIBEYE (magenta) do not seem to impact inactivation of CaV1.3 channels.

B A power exponential function was fitted to the first 5 ms of current traces (as shown in Figure 5B) to obtain the activation time constant (mean ± SEM) at different voltages, which do not appear to be notably impacted.

Larger synthetic ribbon-type AZs establish larger CaV1.3 Ca2+ channel clusters

Next, we tetra-transfected HEK293 cells with palm-Bassoon, Halo-tagged CaV1.3α1, RBP2 and RIBEYE to explore the impact of co-expressing of RIBEYE on the clustering of CaV1.3 Ca2+ channels (Figure 6A for transfection scheme). Using confocal imaging of immunofluorescently labelled RIBEYE, Halo-CaV1.3 and Bassoon (RBP2 immunofluorescence was imaged using epifluorescence to select tetra-transfected HEK293 cells), we observed colocalization of CaV1.3α1, Bassoon and RIBEYE (Fig. 6B). Line profiles drawn tangentially along the plasma membrane showed that beyond the general distribution of a CaV1.3α1 signal in the plasma membrane, CaV1.3α1 signal hotspots occurred underneath the SyRibbons (Fig. 6C). The signal intensity of RIBEYE immunofluorescence positively correlated with the CaV1.3 signal immunofluorescence, (Pr = 0.839 and 0.886 from two sample line scans, ****P < 0.0001), which has previously been reported for immunolabeled native ribbon-type AZs of mouse IHCs (Ohn et al, 2016). We next turned to 2-color STED imaging which provided improved spatial resolution and highlighted the confined localization of CaV1.3 Ca2+ -channel clusters underneath SyRibbons as shown in Figure 6D.

Synthetic ribbon-type active zones recruit CaV1.3 Ca2+ channels.

A Experimental scheme for expression of Halo tagged CaV1.3, palm-Bassoon, RIBEYE-GFP and RBP2 in HEK293 cells.

B Representative confocal images (single sections) of HEK293 cells transfected with RBP2 (not shown, expression confirmed using epifluorescence), RIBEYE-GFP (red), palm-Bassoon (green), Halo-CaV1.3 (blue) shows colocalization of the three latter proteins at the plasma membrane. Scale bar = 5 µm.

C Confocal zoom-ins from (B). Line scans depict an increased CaV1.3 signal intensity (blue) at sites where RIBEYE (red) and palm-Bassoon (green) clusters localize. Pearson’s correlation coefficients (Pr) were calculated along the line profiles for RIBEYE and CaV1.3 signal intensity and indicate a high degree of correlation (∼0.8) between the spatial localization of the two. Scale bar = 2 µm.

D Super-resolution STED images showing CaV1.3 clusters localizing at the base of the SyRibbons. Scale bar = 500nm.

E Distribution of volumes from confocal images of CaV1.3 clusters colocalizing with SyRibbons (orange, nspots = 634) and not colocalizing with SyRibbons (yellow, nspots = 1354) from HEK293 cells expressing CaV1.3, RBP2 and SyRibbons (N = 13 cells). Pooled volumes of all CaV1.3 clusters (with and without SyRibbons) from these cells have been shown in red (nspots = 1988). Volumes of CaV1.3 clusters from HEK293 cells expressing only Halo-CaV1.3 (black, nspots = 1281, N = 16 cells) have been plotted for comparison.

F Box plot depicting data from (E). CaV1.3 clusters appear larger on an average in cells expressing CaV1.3, RBP2 and SyRibbons versus in cells expressing only CaV1.3, (**P = 0.0044). Within cells expressing all four proteins, CaV1.3 clusters are much larger when colocalizing with SyRibbons versus when they do not (****P < 0.0001,). Finally, when comparing these clusters with and without SyRibbons in cells expressing all four proteins versus clusters in cells expressing only CaV1.3, we observe a significant difference for the former (****P < 0.0001,), but not the latter (P = 0.099,). Statistical test: Kruskal-Wallis test with post hoc Dunn’s multiple comparison correction. Overlaid plus signs represent individual spots, crosses represent mean values, central band indicates the median, whiskers represent 90/10 percentiles and boxes represent 75/25 percentiles.

We performed surface renderings of confocally imaged Halo-tagged CaV1.3α1, RIBEYE and palm-Bassoon immunofluorescence spots as described above for estimating the volume of SyRibbons and the CaV1.3 Ca2+ channel clusters (Fig.6E, F). This revealed significantly larger CaV1.3 Ca2+ channel clusters when colocalizing with SyRibbons in HEK293 + Halo-CaV1.3 + RBP2 + SyRibbons cells (average volume ± SD = 0.18 ± 0.28 µm3, nspots= 634, N = 13 cells, data representative of 5 transfections) compared to non-colocalized CaV1.3 Ca2+ channel clusters from the same cells (volume = 0.06 ± 0.10 µm3, nspots= 1354, N = 13 cells; ****P < 0.0001, Kruskal-Wallis test with post hoc Dunn’s multiple comparison test) and CaV1.3 Ca2+ clusters in HEK293 cells solely expressing Halo-CaV1.3 Ca2+ channels (volume = 0.07 ± 0.09µm3, nspots= 1281, N = 16 cells, data representative of 4 transfections; ****P < 0.0001, Kruskal-Wallis test with post hoc Dunn’s multiple comparison test). CaV1.3 Ca2+ channel cluster volumes in the latter two did not differ significantly (P = 0.099, Kruskal-Wallis test with post hoc Dunn’s multiple comparison test). Moreover, volumes of CaV1.3 cluster surfaces show a moderate positive correlation with volumes of corresponding RIBEYE surfaces (Pr = 0.385, ****P < 0.0001), similar to previous observations at IHC ribbon synapses (Frank et al, 2009; Michanski et al, 2023; Ohn et al, 2016).

Ca2+ imaging reveals increased Ca2+ signal intensity at synthetic ribbon-type active zones

Finally, we set-out to functionally characterize synthetic ribbon-type AZs. We combined patch-clamp recordings of CaV1.3 Ca2+ influx with spinning-disk confocal microscopy to visualize Ca2+ signals at the synthetic ribbon-type AZs in HEK293 cells. Following tetracycline-induction for CaV1.3 expression, HEK293 cells were co-transfected with constructs expressing RIBEYE-GFP, untagged palm-Bassoon and RBP2-p2A-mKATE2, (Figure 7A). We recorded mKATE2 positive cells that expressed peripheral GFP puncta, indicative of RIBEYE and palm-Bassoon co-expression. We employed peripheral GFP expression as a proxy of SyRibbons for functional analysis (Figure 7B). Ruptured patch-clamp recordings were performed with 10 mM [Ca2+]e and 10 mM intracellular EGTA to enhance the signal to background ratio for visualizing Ca2+ influx using the low-affinity red-shifted Ca2+ indicator Calbryte590 (100 µM, kd = 1.4 µM). After loading the cell for approximately a minute, we first applied step depolarizations of 20 ms from −83 to +62 mV in 5-mV step increments to measure Ca2+ current (density)–voltage relations.

Ca2+ imaging reveals higher Ca2+ signal intensity underneath SyRibbons.

A Experimental scheme for expression of SyRibbons, RBP2 and CaV1.3 in HEK293 cells for patch clamp in combination with Ca2+ imaging using the low affinity Ca2+ indicator Calbryte590.

B Exemplary cell used for Ca2+ imaging. Cells were identified by peripheral RIBEYE-GFP puncta as shown (indicative of RIBEYE and palm-Bassoon co-expression) and mKATE2 signal (indicative of RBP2 expression).

C Current density-voltage (IV) relations from whole-cell patch clamp recordings of HEK293 cells expressing CaV1.3 + RBP2 + SyRibbons (n=20, in orange) and only CaV1.3 (n=10, in black). Lines represent mean current traces, shaded area represents ± SEM. Whole-cell Ca2+ current density does not appear to change upon co-expression of RBP2 and SyRibbons (P = 0.293, t-test).

D The voltage-dependence of Ca2+ current influx does not seem to be altered upon co-expression of RBP2 and SyRibbons at the whole-cell level.

E A depolarizing pulse to +2mV was applied to the cells for 500 ms and the increase in Calbryte590 fluorescence was measured by acquiring images at a frame rate of 20 Hz using a spinning disk confocal microscope.

F Representative image of a HEK293 cell expressing SyRibbons, RBP2 and CaV1.3 stimulated as described in (E). RIBEYE-GFP signal has been shown in magenta; the change in Calbryte590 fluorescence intensity (ΔF) upon Ca2+ binding (the frame at the onset of depolarization to +2mV) has been shown in green. Note the distinct Ca2+ signal “hotspots” that co-localize with SyRibbons. Scale bar = 10 µm.

G Zoom-ins from (F). Frames before, at the onset of, and after stimulation have been shown. Note the localized Ca2+ influx at the base of the SyRibbons. Scale bar = 1 µm.

H Line profiles drawn tangentially to the membrane in composite ΔF image of Calbryte590 signal and RIBEYE-GFP show a high correlation between the localization of SyRibbons and peak intensity of Calbryte590 fluorescence increase (Pr = ∼0.7). Plots show intensity profiles along line scans for RIBEYE-GFP (magenta) and Calbryte590 (light grey for frames before stimulus, bold green for peak intensity at onset of stimulus, light green lines for decaying intensity during ongoing stimulus and dark grey for frames post-stimulus).

I Regions of interest (ROIs) of 2 µm diameter were drawn at sites with (magenta circles) and without (green circles) SyRibbons and the corresponding Calbryte590 ΔF/F0 values were calculated for each ROI.

J Plot of average (shaded area represents ± SEM) ΔF/F0 values from ROIs with and without SyRibbons (218 and 139 ROIs respectively from N = 24 cells, data from 7 transfections).

K On average, Calbryte590 ΔFmax/F0 was higher for ROIs with SyRibbons than ROIs without them in a given cell (*P = 0.018, paired t-test). Dots represent mean of ΔFmax/F0 from ROIs with (magenta) and without (green) SyRibbons averaged per cell, error bars represent ± SEM.

We did not observe any changes in either the Ca2+ current density (P = 0.293, t-test) or the voltage-dependence of Ca2+ current (P = 0.780, t-test when comparing Vhalf) when comparing cells expressing CaV1.3 + RBP2 + SyRibbons with cells expressing only CaV1.3 (Figure 7C and 7D). Next, we applied a depolarizing pulse to +2 mV for 500 ms and acquired images at a frame rate of 20 Hz (Figure 7E). The intracellular Ca2+ signals mediated by CaV1.3 Ca2+ influx were visualized as an increase in Calbryte590 fluorescence. The increase of Calbryte590 fluorescence appeared to spread throughout the HEK293 cell membrane with discrete regions of higher signal intensity at sites underneath the SyRibbons (exemplary cell shown in Figure 7F with zoom-in at site with SyRibbon shown in 7G). We note that we did not find noticeable red fluorescence at SyRibbons in the absence of depolarization, which is similar to findings with Fluo-4FF in IHCs (Frank et al, 2009), but different from the situation for the large spherical presynaptic bodies in bullfrog hair cells that are stained by Fluo-3 (Issa & Hudspeth, 1996). Line profiles drawn tangentially to the membrane in composite ΔF image of Calbryte590 signal and RIBEYE-GFP showed a high correlation between the localization of SyRibbons and peak intensity of Calbryte590 fluorescence increase, indicating preferential Ca2+ signaling at the SyRibbons (two representative line scans shown in Figure 7H, Pearson correlation coefficients of 0.783 and 0.757). Recordings were made from 24 cells (7 transfections) and ROIs (diameter = 2 µm) at regions with and without SyRibbons were analyzed as shown in Figure 7I. Calbryte590 ΔFmax/F0 values were calculated for each ROI which on an average showed higher ΔFmax/F0 values for ROIs with SyRibbons (on average ∼26% higher) than for ROIs without SyRibbons (*P = 0.018, paired t-test, Figure 7J). We note that occasionally in some cells we observed a higher Calbryte590 signal increase at sites without SyRibbons, and we attribute this to the likely presence of AZ-like clusters composed of RBP2 and palm-Bassoon in these regions which we could not visualize.

Discussion

In this study, we reconstituted and characterized a minimal ribbon-type AZ model system using heterologous expression. Co-expressing CaV1.3 Ca2+ channels with membrane targeted Bassoon for RIBEYE anchorage to the plasma membrane and attracting exogenous RBP2 for clustering Ca2+ channels in HEK293 cells led to structures recapitulating basic aspects of IHC ribbon synapses. Despite the poor regulation of transgene expression in the synapse-naïve HEK293 cells, the AZ-like structures, in the subset of cells were similar in morphology to the native IHC AZs. This applied to the synthetic ribbons as well as to the AZ-like clusters of Bassoon, RBP2 and CaV1.3 Ca2+ channels. However, all three components exhibited variability beyond the substantial natural heterogeneity found among the IHC AZs (Moser et al, 2023). Functionally, Bassoon and RBP2 co-expression resulted in larger whole-cell CaV1.3 currents. While additional expression of RIBEYE does not seem to enhance whole-cell Ca2+ currents, we demonstrate a stronger, localized Ca2+ influx at SyRibbons. As SyRibbons partially resemble native hair cell AZs, we expect this easily available and experimentally accessible system to serve as an advanced testbed for functional interactions of AZ proteins and Ca2+ channels. Further efforts towards reconstituting synaptic vesicles and their release sites will help to enhance the utility of these synthetic AZs.

Synthetic active zone reconstitution model to study ribbon synapse assembly

By demonstrating that membrane-targeted Bassoon suffices to anchor ribbon-like RIBEYE assemblies, this work adds to the top-down and bottom-up evidence for a key role of Bassoon in attracting RIBEYE to the presynaptic density at the AZ in photoreceptors and hair cells (Dick et al, 2003; Khimich et al, 2005; Jing et al, 2013; tom Dieck et al, 2005). Additional expression of multi-domain proteins of the presynaptic density such as CAST (Ohtsuka et al, 2002; Inoue et al, 2006) might replace the need for artificial palmitoylation of Bassoon for membrane targeting. Furthermore, our results support a model of a bidirectional control of AZ size and shape by RIBEYE and proteins of the presynaptic density. The amount of RIBEYE assembled in membrane localized SyRibbons appeared regulated strongly contrasting the seemingly unregulated cytosolic RIBEYE assemblies dominating HEK293 cells in the absence of membrane-targeted Bassoon. Vice versa, clusters of Bassoon at the plasma membrane were larger at SyRibbons than in the absence of RIBEYE. Finally, the size of SyRibbons seemingly scaled with the size of CaV1.3 and Bassoon clusters similar to what is observed in IHCs (Ohn et al, 2016). Such a model is consistent with results from genetic perturbation of native ribbon synapses: i) disruption of RIBEYE leads to a disintegration of Bassoon into smaller subclusters in IHCs (Jean et al, 2018), and ii) disruption of CAST and ELKS reduces the size of the ribbon-type AZs in rod photoreceptors (tom Dieck et al, 2012; Hagiwara et al, 2018). We speculate that achieving the full extent of large ribbon-type AZs such as in rod photoreceptors as well as regulating the size of ribbon-type AZs at a specific synapse according to the precise functional demands requires such a functional interplay between proteins of the presynaptic density and RIBEYE. We note that other ribbon-resident proteins such as piccolino likely contribute to this fine-tuning of the size and shape of ribbon-type AZs. Indeed, disruption of piccolino in IHCs reduced the average ribbon size (Müller et al, 2019; Michanski et al, 2023). One exciting example of ribbon synapse heterogeneity manifests itself in IHCs where synapses show a spatial size gradient that likely relates to the diverse molecular and functional properties of the postsynaptic SGNs (reviewed in Moser et al, 2023).

Insights into presynaptic Ca2+ channel clustering and function

Current discussion of the clustering of CaV Ca2+ channels offers two different models: low affinity protein-protein interactions (e.g. Hibino et al, 2002; Kaeser et al, 2011) via specific domains and liquid-liquid phase separation involving intrinsically disordered domains (e.g. Heck et al, 2019). Work on ribbon synapses has considered several layers of organizing CaV Ca2+ channels at the AZ that seem more compatible with the former model: i) macro-clustering to which the synaptic ribbon contributes as a “super-scaffold” (Jean et al, 2018; Neef et al, 2018; Frank et al, 2010; Maxeiner et al, 2016), ii) tethering and micro-clustering by RBPs, RIMs, CAST and ELKS (Liu et al, 2011; Grabner et al, 2015; Jung et al, 2015; Krinner et al, 2017; Luo et al, 2017; Hagiwara et al, 2018) and iii) nanoscale coupling of CaV Ca2+ channels and SV release sites (Brandt, 2005; Jarsky et al, 2010; Wong et al, 2014; Maxeiner et al, 2016; Özçete & Moser, 2021; Grabner & Moser, 2021; Jaime Tobón & Moser, 2023) for which a role of the ribbon has been more controversial (e.g. Maxeiner et al, 2016; Grabner & Moser, 2021). MINFLUX nanoscopy of rod photoreceptor AZs recently revealed a well-ordered double-line array topography of CaV Ca2+ channels, RIM, Bassoon and ubMunc13-2 at the presynaptic membrane on both sides of the ribbon (Grabner et al, 2022). Clearly, we could not fully reconstitute this complex organization with the minimal set of molecular players at syRibbons in HEK293 cells. Yet, we demonstrate augmentation of CaV1.3 Ca2+ currents and clusters by co-expression of RBP and Bassoon and a positive effect of syRibbons on local Ca2+ signaling, recapitulating micro- and macro-clustering of CaV1.3 Ca2+ channels. However, we note that the functional impact of ribbon-mediated macro-clustering of CaV1.3 Ca2+ channels remains unclear at the moment. Our Ca2+ imaging data seems more reminiscent of previous work on ribbon-type AZs of immature IHCs (Wong et al, 2014) where a broader distribution of CaV1.3 and ribbons go along with an immature state of the IHC AZs. One clear limitation of the synthetic AZ system in HEK293 cells is the lack of SVs and the molecular machinery required for SV release. Nonetheless, our work paves the way for future studies including stable co-expression of AZ proteins including of the release site marker Munc13-1 (Sakamoto et al, 2018; Böhme et al, 2016) and potential delivery of SV machinery via co-expression (Park et al, 2021) or from synaptosomes. Alternatively, using neurosecretory cells such as pheochromocytoma cells (PC12) or adrenal chromaffin cells in primary culture, offer the advantage of providing synaptic-like microvesicles and a large set of components of the presynaptic machinery such that expression of exogenous proteins might be limited to RIBEYE. On the other hand, the complex molecular background involving various types of CaV channels will likely complicate the interpretation. Regardless, SyRibbons in cultured cells offer great availability and as adherent cells, are well accessible to sophisticated analysis by techniques such MINFLUX nanoscopy and cryo-electron tomography. Future use of polycistronic gene expression or stable cell lines expressing all components could help make the system become even more widely applicable.

Materials and Methods

HEK cell culture and transfections

In this study, we used human embryonic kidney (HEK293) cells stably expressing a tetracycline-inducible human CaV1.3 pore forming α1-D subunit transgene (CACNA1D, NM_000720.2), and showing constitutive expression of CaVβ3 (CACNB3, NM_000725.2) and CaVα2δ-1 (CACNA2D1, NM_000722.2). The stably expressing cell line was acquired from Charles River Laboratories, Cleveland, Ohio, USA (Cat. No. CT6232). The cells were cultured as per product protocols in Dulbecco’s Modified Eagle Medium (DMEM) containing GlutaMax, high glucose and pyruvate (Gibco, Life Tech., 31966047), supplemented with 10% Fetal Bovine Serum (FBS; Gibco, Life Tech., A5256701) and 100 units/ml of Penicillin-Streptomycin (Gibco, Life Tech., 15070063). Additionally, the media was supplemented with 0.6 µM of Isradipine (Sigma Aldrich, I6658) and the following selection antibiotics (in mg/ml): 0.005 Blasticidin (Invivogen, ant-bl-05), 0.50 Geneticin (G418 Sulfate; Gibco, Life Tech., 10131027), 0.10 Zeocin (InvivoGen, ant-zn-05) and 0.04 Hygromycin (Thermo Fisher, 10687010). Cells were cultured in a humidified incubator at 37°C with 5% (v/v) CO2 saturation. Cells were split at ∼70% confluency every 3-4 days to prevent adverse effects on cell growth and channel expression, by dissociating the cell using Accutase (Sigma-Aldrich, A6964). The passage number did not exceed more than 26 passages. For experiments, cells were grown in media lacking selection antibiotics. For induction of α1-D subunit expression, cells were treated with selection antibiotic-free medium containing 3 µg/ml tetracycline. For transfection, 6.8 µg polyethylenimine (PEI, 25kDa linear, Polysciences, 23966) was added along with a total of 2 µg of DNA to a final volume of 100 µl of DMEM (without FBS). The PEI/DNA mixture was thoroughly mixed and allowed to incubate for 30 min at room temperature before being added to the cells (30 µl/well in a 24-well-plate containing 500 µl media, cell confluency ∼50-70%). For all experiments with co-transfections, we used an equimolar ratio of DNA (total amount of DNA was kept 2 µg in a 100 µl transfection mix). After 18-24 hours, transfection media was replaced with fresh media devoid of selection antibiotics. Cells were used for experiments (immunocytochemistry, patch clamp and Ca2+ imaging) 24-48 hours after changing the media or as specified. If cell confluency was too high, cells were reseeded to an appropriate confluency and allowed to settle for at least 12 hours before commencement of experiments.

Expression vectors for RIBEYE, Bassoon, CaV1.3 and RBP2

The RIBEYE-GFP construct used consisted of a human RIBEYE cDNA cloned inframe into a pAAV-GFP vector driven by a hybrid CMV-enhancer-human-beta-actin promoter (CMV-HBA) and containing a downstream Woodchuck Hepatitis Virus (WHV) Posttranscriptional Regulatory Element (WPRE) for mRNA stabilization. All Bassoon constructs encode rat Bassoon. The untagged full-length constructs were cloned in the pCS2+ vector. For the generation of palm-Bassoon constructs, we used a pEGFP-C1 vector backbone with a CMV promoter and kanamycin resistance cassette. The insert comprised of a palmitoylation consensus sequence of GAP43 (ATGCTGTGCTGTATGAGAAGAACCAAACAGGTTGAAAAGAATGATGAGGACCAAAAGATTTCCGGACTCAGA TCTCGAG), followed by a cDNA sequence encoding amino acids 95-3938 of rat Bassoon, and a C-terminal monomeric GFP. For the untagged version, the GFP tag was replaced by a STOP codon. To generate the Halo-CaV1.3 plasmid (Schwenzer et al, 2024), human CaV1.3 cDNA (accession number NM_001128840.2) was de-novo synthesized and assembled into a HaloTag vector (Promega G7721) using restriction cloning, thus encoding an N-terminal fusion of CaV1.3 to HaloTag linked by a ‘GGS’ sequence. The analogous GFP-CaV1.3 plasmid was generated by exchanging the HaloTag sequence for mEGFP using restriction cloning. The RBP2 construct with an mKATE2 tag comprised of an insert coding for mouse RBP2 (accession ID NP_001074857.1) with a p2A cleavage site followed by a C-terminal mKATE2 cloned inframe into a f(syn)w-mKATE2-p2A vector driven by a CMV promoter-enhancer and containing downstream WPRE sequence. The untagged version was made by cloning the RBP2 cDNA from this plasmid and replacing RIBEYE-GFP in the pAAV-CMV-HBA-RIBEYE-GFP-WPRE plasmid by in-fusion cloning (Takara Bio USA, Inc.).

Immunocytochemistry and Imaging

HEK293 cells plated on poly-L-lysine coated coverslips were fixed with 99% chilled methanol at −20°C for 2 minutes as previously described (Picher et al, 2017b). The coverslips were washed thoroughly three times with PBS at room temperature (5 – 10 min). Blocking and permeabilization was performed with GSDB (goat serum dilution buffer: 16% normal goat serum, 450 mM NaCl, 0.3% Triton X-100, 20mM phosphate buffer, pH ∼7.4) for 45-60 minutes at room temperature. Samples were incubated with respective primary antibodies (diluted in GSDB, refer to Table EV2) overnight at 4°C or for 2 hours at RT. The samples were then washed three times (5–10 minutes each) with wash buffer (450 mM NaCl, 0.3% Triton X-100, 20mM phosphate buffer, pH ∼ 7.4). Incubation with appropriate secondary antibodies (also diluted in GSDB refer to Table EV2) was performed for 1 hour at room temperature in a light-protected wet chamber. Lastly, the coverslips were washed three times with wash buffer (5-10 minutes each) and one final time with PBS, before mounting onto glass slides with a drop of fluorescence mounting medium (Mowiol 4-88, Carl Roth). For live cell imaging of cells transfected with Halo-CaV1.3 (Figure 4B, C), cells plated on glass-bottomed dishes were treated with Janelia Fluor 646 HaloTag Ligand (Promega, Cat. No. GA1120) at a final labelling concentration of 200nM (in cell culture media). Cells were incubated with the ligand for ∼60 min at 37°C with 5% (v/v) CO2 saturation, after which the media was replaced with an equal volume of fresh warm culture media.

Images from fixed and live samples were acquired in confocal/ STED mode using an Olympus IX83 inverted microscope combined with an Abberior Instruments Expert Line STED microscope (Abberior Instruments GmbH). We used lasers at 488, 561, and 633 nm for excitation and a 775 nm (1.2W) laser for stimulated emission depletion. 1.4 NA 100X or 20X oil immersion objectives were used for fixed samples, and a water immersion 60X objective was used for live samples. Confocal stacks or single sections were acquired using Imspector Software (pixel size = 80 X 80 nm along xy, 200 nm along z). For 2D-STED images, a pixel size of 15 X 15 nm (in xy) was used. Images in figure EV2A were acquired using a Leica SP8 confocal microscope (Leica Microsystems, Germany). All acquired images were visualized using NIH ImageJ software and adjusted for brightness and contrast. Samples and their corresponding controls were processed in parallel using identical staining protocols, laser excitation powers and microscope settings.

Patch clamp and calcium imaging

CaV1.3 stably expressing HEK293 cells were plated on coverslips till about 40-50% confluency was achieved, following which they were induced for CaV1.3α1-D expression by incubating them for 24-48 hours in induction media supplemented with tetracycline. For transfected cells, transfection was performed after 24hrs of induction in the induction media as explained above. Cells were recorded in extracellular solution containing (in mM): 150 choline-Cl, 10 HEPES, 1 MgCl2 and 10 CaCl2; pH was adjusted to 7.3 with CsOH and osmolality was 300-310 mOsm/kg. For electrophysiological recordings in Figure 5 and Figure S4, pipette solution contained (in mM): 140 NMDG, 10 NaCl, 10 HEPES, 5 EGTA, 3 Mg-ATP and 1 MgCl2. pH was adjusted to 7.3 with methanesulfonic acid and osmolality was around 290 mOsm/kg. For Ca2+ imaging in Figure 7, the pipette solution contained (in mM): 111 Cs-glutamate, 1 MgCl2, 1 CaCl2, 10 EGTA, 13 TEA-Cl, 20 HEPES, 4 Mg-ATP, 0.3 Na-GTP, 1 L-glutathione, 0.1 Calbryte590 (AAT Bioquest, Cat. No. 20706); pH was adjusted to 7.3, osmolality to 290 mOsm/kg. Resistance of patch pipettes was 3–7 MΩ. HEK293 cells were rupture patch clamped with EPC-10 amplifier (HEKA electronics, Germany) controlled by Patchmaster software at room temperature, as described previously (Picher et al, 2017b, 2017a). Cells were kept at a holding potential of −91.2 or −88 mV. All voltages were corrected for liquid junction potential offline (21.2 mV or 18 mV). Currents were leak corrected using a p/10 protocol. Recordings were discarded when leak current exceeded −50 pA, Rs exceeded 15 MΩ or offset potential fluctuated more than 5 mV. Current voltage (IV) relations were recorded around 1 minute after rupturing the cell, by applying increasing step-depolarisation pulses of voltage ranging from −86.2 mV to 58.8 mV, in steps of 5 mV. Depolarisation pulses of 500 ms were used for measure CaV1.3 inactivation in Figure EV4A.

For Ca2+ imaging experiments, we used a Yokogawa CSU-X1A spinning-disk confocal scanner, mounted on an upright microscope (Zeiss Examiner) with a 63X, 1.0 NA objective (W Plan-Apochromat, Zeiss). Images were acquired using an Andor Zyla sCMOS 4.2 camera, controlled by VisiView 5.0 software (Visitron Systems GmbH). GFP, mKATE2 and the low-affinity Ca2+ indicator Calbryte590 were excited by a VS Laser Merge System (Visitron Systems) with 488 and 561 nm laser lines. Spinning disk was set to 5000 rpm. mKATE2 positive cells showing peripheral GFP expression (indicative of RIBEYE and palm-Bassoon co-expression) were identified and used for recordings. The cell was loaded for approximately a minute with Calbryte590 dye, after which we applied step depolarizations of 20 ms from −83 to +62 mV in 5-mV step increments to measure Ca2+ current (density)–voltage relations. Next, a central plane of the cell was selected and the GFP fluorescence (green channel) of the cell showing SyRibbons was imaged. This was immediately followed by imaging of the increments of Calbryte590 fluorescence (red channel) in the same plane triggered by a depolarizing pulse to +2 mV for 500 ms (frame rate = 20 Hz). For Ca2+ imaging experiments, we also included occasionally occurring HEK293 cells which were electrically coupled with other neighbouring HEK293 cells. For such cells, we could not fully compensate for the slow capacitive currents and we did not use them for electrophysiological analysis.

Data analysis

Analysis of confocal images

All images were processed using NIH ImageJ software to make z-projections and multi-channel composites, and figures were created using Adobe Illustrator. For analysis of membrane distribution, a region of interest (ROI) of 1 µm thickness was drawn along the periphery of the cell which was identified either using Na, K ATPase α1 immunofluorescence or by enhancing the contrast in images of transfected cells so that the cell boundary becomes distinguishable. Mean pixel intensity was measured along this ROI and for the area enclosed within the ROI. The ratio of the mean pixel intensity (periphery: inside) per cell was reported as an average from at least three central planes; we did not use the basal and top plane. For volumetric fits of SyRibbons, IHC ribbons and Ca2+ channel clusters, we used the inbuilt surface detection algorithm from Imaris 9.6 (Oxford Instruments). The surface detail was set to 0.16 µm and largest sphere diameters for background subtraction were set as follows: for RIBEYE/CtBP2 0.3 µm, for Bassoon 0.1 µm and for CaV1.3 0.08 µm. Thresholding was performed based on the quality of immunofluorescence to ensure all discernible surfaces were detected. All surfaces with less than 10 voxels were filtered out. A manual check was done to include undetected surfaces, remove surfaces not localizing within cell of interest, and split surfaces that would occasionally be clubbed together. For analysis of SyRibbons, only RIBEYE surfaces colocalizing with palm-Bassoon surfaces were used for quantifications (at 0.06 µm or less). Occasionally we would detect surfaces with volumes larger than 2 µm3 (usually cytosolic) or smaller than 0.02 µm3, which were not considered for analysis. For CaV1.3 cluster volumetric analysis, clusters classified as +SyRibbons were at 0.045 µm or less from palm-Bassoon positive RIBEYE surfaces. All other clusters were classified as –SyRibbons. Only clusters localized at the periphery of the cell were used for analysis.

Analysis of patch clamp and Ca2+ imaging data

Analysis of electrophysiology data and figure preparation was performed using Igor Pro 6 and 7 (WaveMetrics Inc.) and final figures were compiled using Adobe Illustrator. For analyzing current density-voltage relations (IVs), the evoked Ca2+ current was averaged from 5 to 10 ms after depolarization start. Cells expressing current amplitudes less than −49 pA and with current density greater than −60 pA/pF were not taken into account for analysis.

Ca2+ imaging data analysis was performed using a customized script in ImageJ. Briefly, we performed background subtraction in time series with Calbryte590 signal and then averaged 5 frames before stimulation and subtracted these from an average of three frames during stimulation to obtain ΔF images. For visualization, the images were smoothened (3 X 3 unweighted smoothening) and a composite was created by overlaying the ΔF images and the RIBEYE-GFP channel. For analyzing ΔF/F0, we drew circular regions of interest (ROIs) of 2µm diameter at sites with and without SyRibbons in the background subtracted time series. The ΔFmax/F0 was calculated as the of average ΔF/F0 in first three frames of peak Ca2+ signal intensity at stimulation onset in each ROI. To minimize bias while assigning regions as with or without SyRibbons, ROIs were assigned in the RIBEYE-GFP channel, without looking at the corresponding Calbryte590 signal.

Statistics

Data sorting and statistics were performed using MS Excel, Igor Pro 7 and/or GraphPad Prism 10. Numerical data is represented mainly as mean ± standard deviation (SD). The normality of data was assessed using Jarque-Bera test and Kolmogorov-Smirnov test, and the equality of variances was checked using F-test. When comparing two samples, a two-tailed unpaired Student’s t-test (or paired t-test for Figure 7J) was performed for normally distributed samples with equal variances. If the conditions for normality and equality of variances were not met, an unpaired Mann-Whitney-Wilcoxon test was performed. When comparing multiple samples, a one-way ANOVA with post hoc Tukey’s test was used for normally distributed data with equal variances. For samples which were not normally distributed, we used Kruskal-Wallis test with post hoc Dunn’s multi-comparison correction. All box plots are depicted with crosses representing mean values, central bands indicating median, whiskers for 90/10 percentiles, boxes for 75/25 percentiles and individual data points overlaid. Non-significant differences have been indicated as n.s., while significant P values have been depicted as *P for < 0.05, **P < 0.01, ***P < 0.001, and ****P < 0.0001.

eLife Assessment

The authors use a synthetic approach to introduce synaptic ribbon proteins into HEK cells and analyze the ability of the resulting assemblies to cluster calcium channels at the active zone. The use of this ground-up approach is valuable as it establishes a system to study molecular interactions at the active zone. The work relies on a solid combination of super-resolution microscopy and electrophysiology, but would benefit from: (i) additional ultrastructural analysis to establish ribbon formation (in the absence of which the claim of these being synthetic ribbons might not be supported; (ii) data quantification (to confirm colocalization of different proteins); (iii) stronger validation of impact on Ca2+ function; (iv) in depth discussion of problems derived from the use of an over-expression approach.

Reviewer #1 (Public Review):

Summary:

In this manuscript, the authors attempt to reconstitute some active zone properties by introducing synaptic ribbon proteins into HEK cells. This "ground-up" approach can be valuable for assessing the necessity of specific proteins in synaptic function. Here, the authors co-transfect a membrane-targeted bassoon, RBP2, calcium channel subunits and Ribeye to generate what they call "synthetic ribbons". The resultant structures show an ability to cluster calcium channels (Figure 4B) and a modest ability to concentrate calcium entry locations (figure 7J). At the light level, the ribeye aggregates look spherical and localize to the membrane through its interaction with the membrane-targeted bassoon. It is a nice proof-of-principle in establishing a useful experimental system for studying calcium channel localization. However, the impact of the study is modest. No new biology is discovered and to call these structures "synthetic ribbons" is an overstatement in the absence of an ultrastructural analysis.

Strengths:

(1) The authors establish a new experimental system for the study of calcium channel localization to active zones.

(2) The clustering of calcium channels to bassoon via RBP2 is a nice confirmation of a previously described interaction between bassoon and calcium channels in a cell-based system

(3) The "ground-up" approach is an attractive one and theoretically allows one to learn a lot about the essential interactions for building a ribbon structure.

Weaknesses:

(1) Are these truly "synthetic ribbons". The ribbon synapse is traditionally defined by its morphology at the EM level. To what extent these structures recapitulate ribbons is not shown. It has been previously shown that Ribeye forms aggregates on its own. Do these structures look any more ribbon-like than ribeye aggregates in the absence of its binding partners?

(2) No new biology is discovered here. The clustering of channels is accomplished by taking advantage of previously described interactions between RBP2, Ca channels and bassoon. The localization of Ribeye to bassoon takes advantage of a previously described interaction between the two. Even the membrane localization of the complexes required the introduction of a membrane-anchoring motif.

(3) The only thing ribbon-specific about these "syn-ribbons" is the expression of ribeye and ribeye does not seem to participate in the localization of other proteins in these complexes. Bsn, Cav1.3 and RBP2 can be found in other neurons.

(4) As the authors point out, RBP2 is not necessary for some Ca channel clustering in hair cells, yet seems to be essential for clustering to bassoon here.

(5) The difference in Ca imaging between SyRibbons and other locations is extremely subtle.

(6) The effect of the expression of palm-Bsn, RBP2 and the combination of the two on Ca-current is ambiguous. It appears that while the combination is larger than the control, it probably isn't significantly different from either of the other two alone (Fig 5). Moreover, expression of Ribeye + the other two showed no effect on Ca current (Figure 7). Also, why is the IV curve right shifted in Figure 7 vs Figure 5?

(7) While some of the IHC is quantified, some of it is simply shown as single images. EV2, EV3 and Figure 4a in particular (4b looks convincing enough on its own, but could also benefit from a larger sample size and quantification)

Reviewer #2 (Public Review):

Summary:

The authors show that co-expression of bassoon, RIBEYE, Cav1.3-alpha1, Cav-beta3, Cav-alpha2delta1, and RBP2 in a heterologus system (HEK293 cells) is sufficient to generate a protein complex resembling a presyanptic ribbon-type active zone both in morphology and in function (in clustering voltage-gated Ca channels and creating sites for localized Ca2+ entry). If the 3 separate Cav gene products are taken as a single protein (i.e. a Ca channel), the conclusion is that the core of a ribbon synapse comprises 4 proteins: bassoon holds the RIBEYE-containing ribbon to the plasma membrane, and RPB2 binds to bassoon and Ca channels, tethering the Ca channels to the presynaptic active zone.

Strengths:

Good use of a heterologous system with generally appropriate controls provides convincing evidence that a presynaptic ribbon-type active zone (without the ability to support exocytosis), with the ability to support localized Ca2+ entry (a key feature of ribbon-type pre-synapses) can be assembled from a few proteins.

Weaknesses:

(1) Relies on over-expression, which almost certainly diminishes the experimentally-measured parameters (e.g. pre-synapse clustering, localization of Ca2+ entry).

(2) Are HEK cells the best model? HEK cells secrete substances and have a studied-endocytitic pathway, but they do not create neurosecretory vesicles. Why didn't the authors try to reconstitute a ribbon synapse in a cell that makes neurosecretory vesicles like a PC12 cell?

(3) Related to 1 and 2: the Ca channel localization observed is significant but not so striking given the presence of Cav protein and measurements of Ca2+ influx distributed across the membrane. Presumably, this is the result of overexpression and an absence of pathways for pre-synaptic targeting of Ca channels. But, still, it was surprising that Ca channel localization was so diffuse. I suppose that the authors tried to reduce the effect of over-expression by using an inducible Cav1.3? Even so, the accessory subunits were constitutively over-expressed.

Reviewer #3 (Public Review):

Summary:

Ribbon synapses are complex molecular assemblies responsible for synaptic vesicle trafficking in sensory cells of the eye and the inner ear. The Ca2+-dependent exocytosis occurs at the active zone (AZ), however, the molecular mechanisms orchestrating the structure and function of the AZs of ribbon synapses are not well understood. To advance in the understanding of those mechanisms, the authors present a novel and interesting experimental strategy pursuing the reconstitution of a minimal active zone of a ribbon synapse within a synapse-naïve cell line: HEK293 cells. The authors have used stably transfected HEK293 cells that express voltage-gated Ca2+ channels subunits (constitutive -CaV beta3 and CaV alpha2 beta1- and inducible CaV1.3 alpha1). They have expressed in those cells several proteins of the ribbon synapse active zone: (1) RIBEYE, (2) a modified version of Bassoon that binds to the plasma membrane through artificial palmitoylation (Palm-Bassoon) and (3) RIM-binding protein 2 (RBP2) to induce the formation of a minimal active zone that they called SyRibbons. The formation of such structures is convincing, however, the evidence of such structures having an impact enhancing Ca2+-currents, as the authors claim, is rather weak in the present version of the study.

Strengths of the study:

(1) The study is carefully carried out using a remarkable combination of (1) superresolution microscopy, to analyze the formation and subcellular distribution of molecular assemblies and (2) functional assessment of voltage-gated Ca2+ channels using patch-clamp recording of Ca2+-currents and fluorometry to correlate Ca2+ influx with the molecular assemblies formed by AZ proteins. The results are of high quality and are in general accompanied of required control experiments.

(2) The method opens new opportunities to further investigate the minimal and basic properties of AZ proteins that are difficult to study using in vivo systems. The cells that operate through ribbon synapses (e.g. photoreceptors and hair cells) are particularly difficult to manipulate, so setting up and validating the use of a heterologous system more suitable for molecular manipulations is highly valuable.

(3) The structures formed by RIBEYE and Palm-Bassoon in HEK293 cells identified by STED nanoscopy are strikingly similar to the AZs of ribbon synapses found in rat inner hair cells (Figure 2).

Weaknesses of the study:

(1) The results obtained in a heterologous system (HEK293 cells) need to be interpreted with caution. They will importantly speed the generation of models and hypothesis that will, however, require in vivo validation.

(2) The authors analyzed the distribution of RIBEYE clusters in different membrane compartments and correctly conclude that RIBEYE clusters are not trapped in any of those compartments, but it is soluble instead. The authors, however, did not carry out a similar analysis for Palm-Bassoon. It is therefore unknown if Palm-Bassoon binds to other membrane compartments besides the plasma membrane. That could occur because in non-neuronal cells GAP43 has been described to be in internal membrane compartments. This should be investigated to document the existence of ectopic internal Synribbons beyond the plasma membrane because it might have implications for interpreting functional data in case Ca2+-channels become part of those internal Synribbons.

(3) The co-expression of RBP2 and Palm-Bassoon induces a rather minor but significant increase in Ca2+-currents (Figure 5). Such an increase does not occur upon expression of (1) Palm-Bassoon alone, (2) RBP2 alone or (3) RIBEYE alone (Figure 5). Intriguingly, the concomitant expression of Palm-Bassoon, RBP2 and RIBEYE does not translate into an increase of Ca2+-currents either (Figure 7).

(4) The authors claim that Ca2+-imaging reveals increased CA2+-signal intensity at synthetic ribbon-type AZs. That claim is a subject of concern because the increase is rather small and it does not correlate with an increase in Ca2+-currents.

Author response:

The following is the authors’ response to the original reviews

Life Assessment

The authors use a synthetic approach to introduce synaptic ribbon proteins into HEK cells and analyze the ability of the resulting assemblies to cluster calcium channels at the active zone. The use of this ground-up approach is valuable as it establishes a system to study molecular interactions at the active zone. The work relies on a solid combination of super-resolution microscopy and electrophysiology, but would benefit from: (i) additional ultrastructural analysis to establish ribbon formation (in the absence of which the claim of these being synthetic ribbons might not be supported; (ii) data quantification (to confirm colocalization of different proteins); (iii) stronger validation of impact on Ca2+ function; (iv) in depth discussion of problems derived from the use of an over-expression approach.

We thank the editors and the reviewers for the constructive comments and appreciation of our work. Please find a detailed point-to-point response below. In response to the critique received, we have now (i) included an ultrastructural analysis of the SyRibbons using correlative light microscopy and cryo-electron tomography, (ii) performed quantifications to confirm the colocalisation of the various proteins, (iii) discussed and carefully rephrased our interpretation of the role of the ribbon in modulating Ca2+ channel function and (iv) discussed concerns regarding the use of an overexpression system.

Public Reviews:

Reviewer #1 (Public Review):

We would like to thank the reviewer for the comments and advice to further improve our manuscript. We have completely overhauled the manuscript taking the suggestions of the reviewer into account.

(1) Are these truly "synthetic ribbons". The ribbon synapse is traditionally defined by its morphology at the EM level. To what extent these structures recapitulate ribbons is not shown. It has been previously shown that Ribeye forms aggregates on its own. Do these structures look any more ribbonlike than ribeye aggregates in the absence of its binding partners?

We thank reviewer 1 for their constructive feedback and critique of the work.

We agree that traditionally, ribbon synapses have always been defined by the distinct morphology observed at the EM level. However, since the discovery of the core-components of ribbons (RIBEYE and Piccolino) confocal and super-resolution imaging of immunofluorescently labelled ribbons have gained importance for analysing ribbon synapses. A correspondence of RIBEYE immunofluorescent structures at the active zone to electron microscopy observations of ribbons has been established in numerous studies (Wong et al, 2014; Michanski et al, 2019, 2023; Maxeiner et al, 2016; Jean et al, 2018) even though direct correlative approaches have yet to be performed to our knowledge. We have now analysed SyRibbons using cryo-correlative electron-light microscopy. We observe that GFPpositive RIBEYE spots corresponded well with electron-dense structures, as is characteristic for synaptic ribbons (Robertis & Franchi, 1956; Smith & Sjöstrand, 1961; Matthews & Fuchs, 2010). We could also observe SyRibbons within 100 nm of the plasma membrane (see Fig. 3). We have now added this qualitative ultrastructural analysis of SyRibbons in the main manuscript (lines 272 - 294, Fig. 3 and Supplementary Fig. 3).

(2) No new biology is discovered here. The clustering of channels is accomplished by taking advantage of previously described interactions between RBP2, Ca channels and bassoon. The localization of Ribeye to bassoon takes advantage of a previously described interaction between the two. Even the membrane localization of the complexes required the introduction of a membraneanchoring motif.

We respectfully disagree with the overall assessment. Our study emphasizes the synthetic establishment of protein assemblies that mimic key aspects of ribbon-type active zone, defining minimum molecular requirements. Numerous previous studies have described the role of the synaptic ribbon in organising the spatial arrangement of Ca2+ channels, regulating their abundance and possibly also modulating their physiological properties (Maxeiner et al, 2016; Frank et al, 2010; Jean et al, 2018; Wong et al, 2014; Grabner & Moser, 2021; Lv et al, 2016). We would like to highlight that there remain major gaps between existing in vitro and in vivo data; for instance, no evidence for direct or indirect interactions between Ca2+ channels and RIBEYE have been demonstrated so far. While we do indeed take advantage of previously known interactions between RIBEYE and Bassoon (tom Dieck et al, 2005); between Bassoon, RBP2 and P/Q-type Ca2+ channels (Davydova et al, 2014); and between RBP2 and Ltype Ca2+ channels (Hibino et al, 2002), our study tries to bridge these gaps by establishing the indirect link between the synaptic ribbon (RIBEYE) and L-type CaV1.3 Ca2+ channels using a bottom-up approach, which has previously just been speculative. Our data shows how even in a synapse-naive heterologous expression system, ribbon synapse components assemble Ca2+ channel clusters and even show a partial localisation of Ca2+ signal. Moreover, we argue that the established reconstitution approach provides other interesting insights such as laying ground-up evidence supporting the anchoring of the synaptic ribbon by Bassoon. Finally, we expect that the established system will serve future studies aimed at deciphering the role of putative CaV1.3 or CaV1.4 interacting proteins in regulating Ca2+ channels of ribbon synapses by providing a more realistic Ca2+ channel assembly that has been available in heterologous expression systems used so far. In response to the reviewers comment we have augmented the discussion accordingly.

(3) The only thing ribbon-specific about these "syn-ribbons" is the expression of ribeye and ribeye does not seem to participate in the localization of other proteins in these complexes. Bsn, Cav1.3 and RBP2 can be found in other neurons.

The synaptic ribbon made of RIBEYE is the key molecular difference in the molecular AZ ultrastructure of ribbon synapses in the eye and the ear. We hypothesize the ribbon to act as a superscaffold that enables AZ with large Ca2+ channel assemblies and readily releasable pools. In further support of this hypothesis, the present study on synthetic ribbons shows that CaV1.3 Ca2+ channel clusters are larger in the presence of SyRibbons compared to SyRibbon-less CaV1.3 Ca2+ channel clusters in tetratransfected HEK cells (Ca2+ channels, RBP, membrane-anchored Bassoon, and RIBEYE, Fig. 6). In response to the reviewers comment we now added an analysis of triple-transfected HEK cells (Ca2+ channels, RBP, membrane-anchored Bassoon), in which CaV1.3 Ca2+ channel clusters again are significantly smaller than at the SyRibbons and indistinguishable from SyRibbon-less CaV1.3 Ca2+ channel clusters (Fig. 6E, F).

(4) As the authors point out, RBP2 is not necessary for some Ca channel clustering in hair cells, yet seems to be essential for clustering to bassoon here.

Here we would like to clarify that RBP2 is indeed important in inner hair cells for promoting a larger complement of CaV1.3 and RBP2 KO mice show smaller CaV1.3 channel clusters and reduced whole cell and single-AZ Ca2+ influx amplitudes (Krinner et al, 2017). However, a key point of difference we emphasize on is that even though CaV1.3 clusters appeared smaller, they did not appear broken or fragmented as they do upon genetic perturbation of Bassoon (Frank et al, 2010), RIBEYE (Jean et al, 2018) or Piccolino (Michanski et al, 2023). This highlights how there may be a hierarchy in the spatial assembly of CaV1.3 channels at the inner hair cell ribbon synapse (also described in the discussion section “insights into presynaptic Ca2+ channel clustering and function”) with proteins like RBP2 regulating abundance of CaV1.3 channels at the synapse and organising them into smaller clusters – what we have termed as “nanoclustering”; while Bassoon and RIBEYE may serve as super-scaffolds further organizing these CaV1.3 nanoclusters into “microclusters”. Observations of fragmented Ca2+ channel clusters and broader spread of Ca2+ signal seen upon Ca2+ imaging in RIBEYE and Bassoon mutants (Jean et al, 2018; Frank et al, 2010; Neef et al, 2018), and the absence of such a phenotype in RBP2 mutants (Krinner et al, 2017) may be explained by such a differential role of these proteins in organising Ca2+ channel spatial assembly. The data of the present study on reconstituted ribbon containing AZs are in line with these observations in inner hair cells: RBP2 appears important to tether Ca2+ channels to Bassoon and these AZ-like assemblies are organised to their full extent by the presence of RIBEYE. As mentioned in the response to point 3 of the reviewer, we have now further strengthened this point by adding the analysis of SyRibbon-less CaV1.3 Ca2+ channel clusters in tripletransfected HEK cells (Ca2+ channels, RBP, membrane-anchored Bassoon, Fig. 6E, F). Moreover, we have revised the discussion accordingly.

(5) The difference in Ca imaging between SyRibbons and other locations is extremely subtle.

We agree with the reviewer on the modest increase in Ca2+ signal amplitude seen in the presence of SyRibbons and provide the following reasoning for this observation:

(i) It is plausible that due to the overexpression approach, Ca2+ channels (along with RBP2 and PalmBassoon) still show considerably high expression throughout the membrane even in regions where SyRibbons are not localised. Indeed, this is evident in the images shown in the lower panel in Fig. 6B, where Ca2+ channel immunofluorescence is distributed across the plasma membrane with larger clusters formed underneath SyRibbons (for an opposing scenario, please see the cell in Fig. 6B upper panel with very localised CaV1.3 distribution underneath SyRibbons). This would of course diminish the difference in the Ca2+ signals between membrane regions with and without SyRibbons. We note that while the contrast is greater for native synapses, extrasynaptic Ca2+ channels have been described in numerous studies alone for hair cells (Roberts et al, 1990; Brandt, 2005; Zampini et al, 2010; Wong et al, 2014).

(ii) Nevertheless, we do not expect a remarkably big difference in Ca2+ influx due to the presence of SyRibbons in the first place. Ribbon-less AZs in inner hair cells of RIBEYE KO mice showed normal Ca2+ current amplitudes at the whole-cell and the single-AZ level (Jean et al, 2018). However, it was the spatial spread of the Ca2+ signal at the single-AZ level which appeared to be broader and more diffuse in these mutants in the absence of the ribbon, in contrast to the more confined Ca2+ hotspots seen in the wild-type controls.

So, in agreement with these published observations – it appears that presence of SyRibbons helps in spatially confining the Ca2+ signal by super scaffolding nanoclusters into microclusters (see also our response to points 3 and 4 of the reviewer): this is evident from seeing some spatial confinement of Ca2+ signals near SyRibbons on top of the diffuse Ca2+ signal across the rest of the membrane as a result of overexpression in HEK cells.

We have now carefully rephrased our interpretation throughout the manuscript and added further explanation in the discussion section.

(6) The effect of the expression of palm-Bsn, RBP2 and the combination of the two on Ca-current is ambiguous. It appears that while the combination is larger than the control, it probably isn't significantly different from either of the other two alone (Fig 5). Moreover, expression of Ribeye + the other two showed no effect on Ca current (Figure 7). Also, why is the IV curve right shifted in Figure 7 vs Figure 5?

We agree with the reviewer that co-expression of palm-Bassoon and RBP2 seems to augment Ca2+ currents, while the additional expression of RIBEYE results in no change when compared to wild-type controls. We currently do not have an explanation for this observation and would refrain from making any claims without concrete evidence. As the reviewer also correctly pointed out, while the expression of the combination of palm-Bassoon and RBP2 raises Ca2+ currents, current amplitudes are not significantly different when compared to the individual expression of the two proteins (P > 0.05, Kruskal-Wallis test). In light of this, we have now carefully rephrased our MS. Moreover, we would like to thank reviewer 1 for pointing out the right shift in the IV curve which was due to an error in the values plotted on the x-axis. This has been corrected in the updated version of the manuscript.

(7) While some of the IHC is quantified, some of it is simply shown as single images. EV2, EV3 and Figure 4a in particular (4b looks convincing enough on its own, but could also benefit from a larger sample size and quantification)

We have now added quantifications for the colocalisations of the various transfection combinations depicted in the above-mentioned figures collectively in Supplementary Figure 7 and added the corresponding results and methods accordingly.

Reviewer #2 (Public Review):

We would like to thank the reviewer for the comments and advice to further improve our manuscript.

(1) Relies on over-expression, which almost certainly diminishes the experimentally-measured parameters (e.g. pre-synapse clustering, localization of Ca2+ entry).

We acknowledge this limitation highlighted by the reviewer arising from the use of an overexpression system and have carefully rephrased our interpretation and discussed possible caveats in the discussion section.

(2) Are HEK cells the best model? HEK cells secrete substances and have a studied-endocytitic pathway, but they do not create neurosecretory vesicles. Why didn't the authors try to reconstitute a ribbon synapse in a cell that makes neurosecretory vesicles like a PC12 cell?

This is a valid point for discussion that we also had here extensively. We indeed did consider pheochromocytoma cells (PC12 cells) for reconstitution of ribbon-type AZs and also performed initial experiments with these in the initial stages of the project. PC12 cells offer the advantage of providing synaptic-like microvesicles and also endogenously express several components of the presynaptic machinery such as Bassoon, RIM2, ELKS etc (Inoue et al, 2006) such that overexpression of exogenous AZ proteins would have to be limited to RIBEYE only.

However, a major drawback of PC12 cells as a model is the complex molecular background of these cells. We have also briefly described this in the discussion section (line 615 – 619). Naïve, undifferentiated PC12 cells show highly heterogeneous expression of various CaV channel types (Janigro et al, 1989); however, CaV1.3, the predominant type in ribbon synapses of the ear, does not seem to express in these cells (Liu et al, 1996). Furthermore, our attempts at performing immunostainings against CaV1.3 and at overexpressing CaV1.3 in PC12 cells did not prove successful and we decided on refraining from pursuing this further (data not shown).

On the contrary, HEK293 cells being “synapse-naïve” provide the advantage of serving as a “blank canvas” for performing such reconstitutions, e.g. they lack voltage-gated Ca2+ channels and multidomain proteins of the active zone. Moreover, an important practical aspect for our choice was the availability of the HEK293 cell line with stable (and inducible) expression of the CaV1.3 Ca2+ channel complex. Finally, as described in lines 613 – 614 of the discussion section, even though HEK293 cells lack SVs and the molecular machinery required for their release, our work paves way for future studies which could employ delivery of SV machinery via co-expression (Park et al, 2021) which could then be analyzed by the correlative light and electron microscopy workflow we worked out and added during revision.

(3) Related to 1 and 2: the Ca channel localization observed is significant but not so striking given the presence of Cav protein and measurements of Ca2+ influx distributed across the membrane. Presumably, this is the result of overexpression and an absence of pathways for pre-synaptic targeting of Ca channels. But, still, it was surprising that Ca channel localization was so diffuse. I suppose that the authors tried to reduce the effect of over-expression by using an inducible Cav1.3? Even so, the accessory subunits were constitutively over-expressed.

We agree with the reviewer on the modest increase in Ca2+ signal amplitude seen in the presence of SyRibbons. Yes, we employed inducible expression of the CaV1.3a subunit and tried to reduce the effect of overexpression by testing different induction times. However, we did not observe any major differences in expression and observed large variability in CaV1.3 expression across cells irrespective of induction duration. At all time points, there were cells with diffuse CaV1.3 localisation also in regions without SyRibbons which likely reduced the contrast of the Ca2+ signal we observe. We provide the following reasoning for this observation:

(i) It is plausible that due to the overexpression approach, Ca2+ channels (along with RBP2 and PalmBassoon) still show considerable expression along the membrane also in regions where SyRibbons are not localised. Indeed, this is evident in the images shown in the lower panel in Fig. 6B where Ca2+ channel immunofluorescence is distributed across the plasma membrane with larger clusters formed underneath SyRibbons. This would of course diminish the difference in the Ca2+ signals between membrane regions with and without SyRibbons. We note that while the contrast is greater for native synapses, extrasynaptic Ca2+ channels have been described in numerous studies alone for hair cells (Roberts et al, 1990; Brandt, 2005; Zampini et al, 2010; Wong et al, 2014).

(ii) Nevertheless, we do not expect a striking difference in Ca2+ influx amplitude due to the presence of SyRibbons in the first place. Ribbon-less AZs in inner hair cells of RIBEYE KO mice showed normal Ca2+ current amplitudes at the whole-cell and the single-AZ level (Jean et al, 2018). Instead, it was the spatial spread of the Ca2+ signal at the single-AZ level which appeared to be broader and more diffuse in these mutants in the absence of the ribbon, in contrast to the more confined Ca2+ hotspots seen in the wildtype controls.

So, in agreement with these published observations – it appears that presence of SyRibbons helps in spatially confining the Ca2+ signal by super scaffolding nanoclusters into microclusters: this is evident from seeing some spatial confinement of Ca2+ signals near SyRibbons on top of the diffuse Ca2+ signal across the rest of the membrane as a result of overexpression in HEK cells.

We have now carefully rephrased our interpretation throughout the manuscript and added further explanation in the discussion section.

Reviewer #3 (Public Review):

We would like to thank the reviewer for the comments and advice to further improve our manuscript.

(1) The results obtained in a heterologous system (HEK293 cells) need to be interpreted with caution. They will importantly speed the generation of models and hypothesis that will, however, require in vivo validation.

We acknowledge this limitation highlighted by Reviewer 3 arising from the use of an overexpression system and have carefully rephrased our interpretation and discussed possible caveats in the discussion section. We employed inducible expression of the CaV1.3a subunit and tried to reduce the effect of overexpression by testing different induction times. However, we did not observe any major differences in expression and observed large variability in CaV1.3 expression across cells irrespective of induction duration. At all time points, there were cells with diffuse CaV1.3 localisation, even in regions without SyRibbons and this could reduce the contrast of the Ca2+ signal we observe. We provide the following reasoning for this observation:

(i) It is plausible that due to the overexpression approach, Ca2+ channels (along with RBP2 and PalmBassoon) still show considerable expression along the membrane also in regions where SyRibbons are not localised. Indeed, this is evident in the images shown in the lower panel in Fig. 6B where Ca2+ channel immunofluorescence is distributed across the plasma membrane with larger clusters formed underneath SyRibbons. This would of course diminish the difference in the Ca2+ signals between membrane regions with and without SyRibbons. We note that while the contrast is greater for native synapses, extrasynaptic Ca2+ channels have been described in numerous studies alone for hair cells (Roberts et al, 1990; Brandt, 2005; Zampini et al, 2010; Wong et al, 2014).

(ii) Nevertheless, we do not expect a striking difference in Ca2+ influx amplitude due to the presence of SyRibbons in the first place. Ribbon-less AZs in inner hair cells of RIBEYE KO mice showed normal Ca2+ current amplitudes at the whole-cell and the single-AZ level (Jean et al, 2018). Instead, it was the spatial spread of the Ca2+ signal at the single-AZ level which appeared to be broader and more diffuse in these mutants in the absence of the ribbon, in contrast to the more confined Ca2+ hotspots seen in the wildtype controls.

So, in agreement with these published observations – it appears that presence of SyRibbons helps in spatially confining the Ca2+ signal by super scaffolding nanoclusters into microclusters: this is evident from seeing some spatial confinement of Ca2+ signals near SyRibbons on top of the diffuse Ca2+ signal across the rest of the membrane as a result of overexpression in HEK cells.

(2) The authors analyzed the distribution of RIBEYE clusters in different membrane compartments and correctly conclude that RIBEYE clusters are not trapped in any of those compartments, but it is soluble instead. The authors, however, did not carry out a similar analysis for Palm-Bassoon. It is therefore unknown if Palm-Bassoon binds to other membrane compartments besides the plasma membrane. That could occur because in non-neuronal cells GAP43 has been described to be in internal membrane compartments. This should be investigated to document the existence of ectopic internal Synribbons beyond the plasma membrane because it might have implications for interpreting functional data in case Ca2+-channels become part of those internal Synribbons.

In response to this valid concern, we have now included the suggested experiment in Supplementary Figure 1. We investigated the subcellular localisation of Palm-Bassoon and did not find Palm-Bassoon puncta to colocalise with ER, Golgi, or lysosomal markers, suggesting against a possible binding with membrane compartments inside the cell. We have added the following sentence in the results section, line 145 : “Palm-Bassoon does not appear to localize in the ER, Golgi apparatus or lysosomes (Supplementary Fig 1 D, E and F).”

(3) The co-expression of RBP2 and Palm-Bassoon induces a rather minor but significant increase in Ca2+-currents (Figure 5). Such an increase does not occur upon expression of (1) Palm-Bassoon alone, (2) RBP2 alone or (3) RIBEYE alone (Figure 5). Intriguingly, the concomitant expression of PalmBassoon, RBP2 and RIBEYE does not translate into an increase of Ca2+-currents either (Figure 7).

We agree with the reviewer that co-expression of palm-Bassoon and RBP2 seems to augment Ca2+ currents, while the additional expression of RIBEYE results in no change when compared to wild-type controls. We currently do not have an explanation for this observation and would refrain from making any claims without concrete evidence. We also highlight that, while the expression of the combination of palm-Bassoon and RBP2 raises Ca2+ currents, current amplitudes are not significantly different when compared to the individual expression of the two proteins (P > 0.05, Kruskal-Wallis test). In light of this, we have now carefully rephrased our MS.

(4) The authors claim that Ca2+-imaging reveals increased CA2+-signal intensity at synthetic ribbontype AZs. That claim is a subject of concern because the increase is rather small and it does not correlate with an increase in Ca2+-currents.

Thanks for the comment: please see our response to your first comment and the lines 585 – 610 in the discussion section.

Recommendations for the authors:

Reviewer #2 (Recommendations For The Authors):

(1) The authors should have a better discussion of problems derived from over-expression.

Done. Please see above.

(2) Ideally, the authors would repeat the study using a secretory cell line, but this is of course not possible. The idea could be brought forth, though.

As described above in our response to the public review of reviewer 2, we have discussed this idea in the discussion section (refer to lines 615 – 619), emphasizing on both the advantages and the limitations of using a secretory cell line (e.g. PC12 cells) instead of HEK293 cells as a model for performing such reconstitutions.

Reviewer #3 (Recommendations For The Authors):

(1) There are several figures in which colocalization between different proteins is studied only displaying images but without any quantitative data. This should be corrected by providing such a quantitative analysis.

We have now added quantifications for the colocalisations of the various transfection combinations depicted in the above-mentioned figures collectively in Supplementary Figure 7 and added the corresponding results and methods accordingly.

(2) The little increase in Ca2+-currents and Ca2+-influx associated to the clustering of Ca2+-channels to Synribbons is a concern. The authors should discuss if such a minor increase (found only when Palm-Bassoon and RBP2 ae co-expressed) would have or not physiological consequences in an actual synapse. They might discuss the comparison of those results and compare with results obtained in genetically modified mice in which Ca2+-currents are affected upon the removal of AZs proteins. On the other hand, they should explain why Ca2+-currents do not increase when the Synribbons are formed by RIBEYE, Palm-Bassoon and RBP2.

Done. Please see above.

(3) The description of the patch-clamp experiments should be enriched by including representative currents. Did the authors measure tail currents?

We would like to thank the reviewer for the valuable suggestion and have now added representative currents to the figures (see Supplementary Figure 5B). We agree with the reviewer on the importance of further characterizing the Ca2+ currents in the presence and absence of SyRibbons by analysis of tail currents for counting the number of Ca2+ channels by non-stationary fluctuation analysis but consider this to be out of scope of the current study and an objective for future studies.

(4) The current displayed in Figure 7 E should be explained better.

Previous studies have shown that Ca2+-binding proteins (CaBPs) compete with Calmodulin to reduce Ca2+-dependent inactivation (CDI) and promote sustained Ca2+ influx in Inner Hair Cells (Cui et al, 2007; Picher et al, 2017). In the absence of CaBPs, CaV1.3-mediated Ca2+ currents show more rapid CDI as in the case here upon heterologous expression in HEK cells ((Koschak et al, 2001), see also Picher et al 2017 where co-expression of CaBP2 with CaV1.3 inhibits CDI in HEK293 cells). The inactivation kinetics of CaV1.3 are also regulated by the subunit composition (Cui et al, 2007) along with the modulation via interaction partners and given the reconstitution here we do not find the currents very surprising.

(5) Is the difference in Ca2+-influx still significantly higher upon the removal of the maximum value measured in positive Syribbons spots (Figure 7, panel K)?

Yes, on removing the maximum value, the P value increases from 0.01 to 0.03 but remains statistically significant.

(6) In summary, although the approach pioneered by the authors is exciting and provides relevant results, there is a major concern regarding the interpretation of the modulation of Ca2+ channels.

We have now carefully rephrased our interpretation on the modulation of Ca2+ channels.

References

Brandt A (2005) Few CaV1.3 Channels Regulate the Exocytosis of a Synaptic Vesicle at the Hair Cell Ribbon Synapse. Journal of Neuroscience 25: 11577–11585

Cui G, Meyer AC, Calin-Jageman I, Neef J, Haeseleer F, Moser T & Lee A (2007) Ca2+-binding proteins tune Ca2+-feedback to Cav1. 3 channels in mouse auditory hair cells. The Journal of Physiology 585: 791–803

Davydova D, Marini C, King C, Klueva J, Bischof F, Romorini S, Montenegro-Venegas C, Heine M, Schneider R, Schröder MS, et al (2014) Bassoon specifically controls presynaptic P/Q-type Ca(2+) channels via RIM-binding protein. Neuron 82: 181–194

tom Dieck S, Altrock WD, Kessels MM, Qualmann B, Regus H, Brauner D, Fejtová A, Bracko O, Gundelfinger ED & Brandstätter JH (2005) Molecular dissection of the photoreceptor ribbon synapse: physical interaction of Bassoon and RIBEYE is essential for the assembly of the ribbon complex. J Cell Biol 168: 825–836

Frank T, Rutherford MA, Strenzke N, Neef A, Pangršič T, Khimich D, Fejtova A, Gundelfinger ED, Liberman MC, Harke B, et al (2010) Bassoon and the synaptic ribbon organize Ca2+ channels and vesicles to add release sites and promote refilling. Neuron 68: 724–738

Grabner CP & Moser T (2021) The mammalian rod synaptic ribbon is essential for Cav channel facilitation and ultrafast synaptic vesicle fusion. eLife 10: e63844

Hibino H, Pironkova R, Onwumere O, Vologodskaia M, Hudspeth AJ & Lesage F (2002) RIM - binding proteins (RBPs) couple Rab3 - interacting molecules (RIMs) to voltage - gated Ca2+ channels. Neuron 34: 411–423

Inoue E, Deguchi-Tawarada M, Takao-Rikitsu E, Inoue M, Kitajima I, Ohtsuka T & Takai Y (2006) ELKS, a protein structurally related to the active zone protein CAST, is involved in Ca2+-dependent exocytosis from PC12 cells. Genes to Cells 11: 659–672

Janigro D, Maccaferri G & Meldolesi J (1989) Calcium channels in undifferentiated PC12 rat pheochromocytoma cells. FEBS Letters 255: 398–400

Jean P, Morena DL de la, Michanski S, Tobón LMJ, Chakrabarti R, Picher MM, Neef J, Jung S, Gültas M, Maxeiner S, et al (2018) The synaptic ribbon is critical for sound encoding at high rates and with temporal precision. Elife 7: e29275

Koschak A, Reimer D, Huber I, Grabner M, Glossmann H, Engel J & Striessnig J (2001) alpha 1D (Cav1.3) subunits can form l-type Ca2+ channels activating at negative voltages. J Biol Chem 276: 22100–22106

Krinner S, Butola T, Jung S, Wichmann C & Moser T (2017) RIM-Binding Protein 2 Promotes a Large Number of CaV1.3 Ca2+-Channels and Contributes to Fast Synaptic Vesicle Replenishment at Hair Cell Active Zones. Front Cell Neurosci 11: 334

Liu H, Felix R, Gurnett CA, De Waard M, Witcher DR & Campbell KP (1996) Expression and Subunit Interaction of Voltage-Dependent Ca2+ Channels in PC12 Cells. J Neurosci 16: 7557–7565

Lv C, Stewart WJ, Akanyeti O, Frederick C, Zhu J, Santos-Sacchi J, Sheets L, Liao JC & Zenisek D (2016) Synaptic Ribbons Require Ribeye for Electron Density, Proper Synaptic Localization, and Recruitment of Calcium Channels. Cell Reports 15: 2784–2795

Matthews G & Fuchs P (2010) The diverse roles of ribbon synapses in sensory neurotransmission. Nat Rev Neurosci 11: 812–822

Maxeiner S, Luo F, Tan A, Schmitz F & Südhof TC (2016) How to make a synaptic ribbon: RIBEYE deletion abolishes ribbons in retinal synapses and disrupts neurotransmitter release. The EMBO Journal 35: 1098–1114

Michanski S, Kapoor R, Steyer AM, Möbius W, Früholz I, Ackermann F, Gültas M, Garner CC, Hamra FK, Neef J, et al (2023) Piccolino is required for ribbon architecture at cochlear inner hair cell synapses and for hearing. EMBO Rep 24: e56702

Michanski S, Smaluch K, Steyer AM, Chakrabarti R, Setz C, Oestreicher D, Fischer C, Möbius W, Moser T, Vogl C, et al (2019) Mapping developmental maturation of inner hair cell ribbon synapses in the apical mouse cochlea. PNAS 116: 6415–6424

Neef J, Urban NT, Ohn T-L, Frank T, Jean P, Hell SW, Willig KI & Moser T (2018) Quantitative optical nanophysiology of Ca2+ signaling at inner hair cell active zones. Nat Commun 9: 290

Park D, Wu Y, Lee S-E, Kim G, Jeong S, Milovanovic D, Camilli PD & Chang S (2021) Cooperative function of synaptophysin and synapsin in the generation of synaptic vesicle-like clusters in non-neuronal cells. Nat Commun 12

Picher MM, Gehrt A, Meese S, Ivanovic A, Predoehl F, Jung S, Schrauwen I, Dragonetti AG, Colombo R, Camp GV, et al (2017) Ca2+-binding protein 2 inhibits Ca2+-channel inactivation in mouse inner hair cells. PNAS 114: E1717–E1726

Robertis ED & Franchi CM (1956) Electron Microscope Observations on Synaptic Vesicles in Synapses of the Retinal Rods and Cones. J Biophys Biochem Cytol 2: 307–318

Roberts WM, Jacobs RA & Hudspeth AJ (1990) Colocalization of ion channels involved in frequency selectivity and synaptic transmission at presynaptic active zones of hair cells. J Neurosci 10: 3664–3684

Smith CA & Sjöstrand FS (1961) A synaptic structure in the hair cells of the guinea pig cochlea. Journal of Ultrastructure Research 5: 184–192

Wong AB, Rutherford MA, Gabrielaitis M, Pangršič T, Göttfert F, Frank T, Michanski S, Hell S, Wolf F, Wichmann C, et al (2014) Developmental refinement of hair cell synapses tightens the coupling of Ca2+ influx to exocytosis. EMBO J 33: 247–264

Zampini V, Johnson SL, Franz C, Lawrence ND, Münkner S, Engel J, Knipper M, Magistretti J, Masetto S & Marcotti W (2010) Elementary properties of CaV1.3 Ca(2+) channels expressed in mouse cochlear inner hair cells. J Physiol 588: 187–199

Version 2

Abstract

Abstract

Encoding of several sensory modalities into neural signals is mediated by ribbon synapses. The synaptic ribbon tethers synaptic vesicles at the presynaptic active zone (AZ) and may act as a super-scaffold organizing AZ topography. Here we employed a synthetic biology approach to reconstitute structures mimicking ribbon-type AZs in HEK293 cells for probing minimal molecular requirements and studying presynaptic Ca2+ channel clustering. Co-expressing a membrane-targeted version of the AZ-protein Bassoon and the ribbon core protein RIBEYE, we observed structures recapitulating basic aspects of ribbon-type AZs, which we call synthetic ribbons or SyRibbons. Super-resolution STED microscopy and cryo-correlative electron tomography revealed SyRibbons were similar to native ribbons at AZs of cochlear inner hair cells in shape and size. SyRibbons with Ca2+ channel clusters formed upon additional expression of CaV1.3 Ca2+ channels and RIM-binding protein 2 (RBP2). CaV1.3 Ca2+ channel clusters associated with SyRibbons were larger than ribbonless CaV1.3 Ca2+ channels clusters and functional analysis by Ca2+-imaging in combination with patch clamp showed partial confinement of the Ca2+ signal at SyRibbons. In summary, we identify Ca2+ channels, RBP, membrane-anchored Bassoon, and RIBEYE as minimal components for reconstituting a basic ribbon-type AZ. SyRibbons might complement animal studies on molecular interactions of AZ proteins.

Introduction

Synapses of sensory receptor cells of the inner ear and the retina are hallmarked by the presence of an electron-dense structure called the synaptic ribbon, which tethers synaptic vesicles (SVs) at the presynaptic active zone (AZ, [Moser et al, 2019; Matthews & Fuchs, 2010]). Proposed functional rolesof the ribbon include (i) its role as a SV replenishment machine for tireless neurotransmission (Bunt, 1971; Holt et al, 2004; Khimich et al, 2005; Frank et al, 2010; Snellman et al, 2011; Graydon et al, 2011; Vaithianathan et al, 2016), (ii) “super-scaffold” regulating the abundance, topography, and function of AZ players such Ca2+ channels and release sites and their tight coupling (Khimich et al, 2005; Frank et al, 2010; Wong et al, 2014; Maxeiner et al, 2016; Jean et al, 2018; Grabner & Moser, 2021), and iii) coordination of SV release (Heidelberger et al, 1994; Glowatzki & Fuchs, 2002; Singer et al, 2004; Edmonds, 2004; Mehta et al, 2013). Yet, our molecular understanding of how these postulated functions are executed by synaptic ribbons is still limited.

Likewise, the clustering of Ca2+ channels at the AZ, how it is varied among different or even the same synapse types and how the ribbon contributes towards this, remain active areas of research (Wichmann & Kuner, 2022; Moser et al, 2023). For example, ribbon synapses of mature inner hair cells (IHC) assemble CaV1.3 voltage-gated Ca2+ channels at the base of the ribbon, but vary in shape, size and channel complement of the CaV1.3 clusters (Frank et al, 2010; Wong et al, 2014; Neef et al, 2018) partially dependent on the position within the IHC (Ohn et al, 2016; Özçete & Moser, 2021). IHC synapses lacking AZ-anchored ribbons in Bassoon mutant mice showed fewer Ca2+ channels with altered topography (Frank et al, 2010; Neef et al, 2018). While genetic disruption of Ca2+ channel-tethering AZ proteins RIM2 (Jung et al, 2015) and RBP2 (Krinner et al, 2017) reduced the abundance of Ca2+ channels at IHC synapses, their topography retained the typical stripe-like shape of Ca2+ channel clusters. Constitutive deletion of RIBEYE, the core scaffold protein of the synaptic ribbon, provided so far the most direct test of the scaffolding role of the ribbon (Maxeiner et al, 2016; Jean et al, 2018; Becker et al, 2018; Grabner & Moser, 2021). Here, ribbon-less synapses of IHCs showed an altered Ca2+ channel topography with a more spatially widespread presynaptic Ca2+ signal but intact Ca2+ channel complement (Jean et al, 2018). Moreover, the voltage-dependence of activation as well as the inactivation of Ca2+ channels appeared to be altered, suggesting a possible role of the synaptic ribbon in regulating Ca2+ channel physiology (Jean et al, 2018). Changes in Ca2+ channel physiology upon deletion of RIBEYE were also found for CaV1.4 at ribbon-less rod photoreceptor AZs (Grabner & Moser, 2021). Finally, IHCs lacking piccolino, a ribbon-synapse specific isoform of the AZ protein piccolo, show AZs with smaller synaptic ribbons and altered clustering, yet normal complement of Ca2+ channels (Michanski et al, 2023).

So how does the synaptic ribbon tune into the orchestra of AZ proteins that collectively organize Ca2+ channel abundance, topography, and function? To what extent can the effects observed upon disruption of Bassoon be attributed to the concomitant loss of the ribbon? These questions seem even more relevant given that, so far to our knowledge, no direct interactions between Bassoon (Frank et al, 2010) or RIBEYE and CaV1.3 Ca2+ channel have been described. Here, we adopted a bottom-up synthetic biology approach to assemble a minimal presynaptic protein machinery required to bring together RIBEYE and Ca2+ channels in a ‘synapse-naïve’ expression system to assess the role of the synaptic ribbon in regulating Ca2+ channel clustering and physiology. Similar approaches have previously been employed for reconstituting aspects of conventional synapses, for instance in “hemi-synapses” between co-cultured neurons and non-neuronal cells overexpressing presynaptic or postsynaptic components (for review see Craig et al, 2006). More recently, Munc13-1 supra-molecular assemblies which recruit release machinery proteins were reconstituted (Sakamoto et al, 2018).

Expression of RIBEYE in synapse-naïve cell lines such as COS-7 cells led to large cytosolic assemblies (Schmitz et al, 2000). Ribbon-like electron dense structures with vesicles have also been reported upon heterologous RIBEYE expression in retinal progenitor cell-line (R28, [Magupalli et al, 2008]). However, these cells might not be considered synapse-naïve and the presence and function of Ca2+ channels at synthetic ribbon synapses remained to be studied. The need for co-expressing at least three Ca2+ channel subunits (CaVαx, CaVβx and CaVα28x) in synapse-naïve cell lines renders acute co-expression of AZ multidomain proteins cumbersome. Here we took advantage of synapse-naïve Human Embryonic Kidney 293 (HEK293) cells stably expressing inducible CaV1.3α1, which is the predominant subtype of voltage-gated Ca2+ channels at IHC ribbon synapses (Brandt et al, 2003; Platzer et al, 2000; Dou et al, 2004), as well as constitutively expressed CaVβ3 and CaVα281. We co-expressed RIBEYE along with membrane targeted Bassoon and observed structures with striking resemblance to IHC synaptic ribbon-type AZs nicknamed ‘SyRibbons’. We characterized the structure and function of these synthetic ribbon-type AZs and identified Bassoon, RBP2, RIBEYE and CaV1.3 channels as the minimal components required for assembling a basic ribbon-type AZ. We observed larger CaV1.3 channel clusters near SyRibbons and with partially localized Ca2+ influx upon stimulation. Our results support the role of synaptic ribbons in promoting the formation of CaV1.3 channel micro-clustering. Synthetic ribbon-type AZs offer a novel approach for functionally studying protein-protein interaction and will likely complement, refine and reduce experiments on native ribbon synapses.

Results

Membrane-targeted Bassoon recruits RIBEYE to the cell membrane in HEK293 cells where structures mimicking inner hair cell synaptic ribbons are formed

HEK293 cells have the advantage of not expressing the synaptic machinery components studied here. This provides a clean background for reconstituting synthetic synapses from a minimal set of synaptic proteins. Our first step towards assembling a ribbon-type AZ in a heterologous expression system was to express RIBEYE, the core scaffold protein of the synaptic ribbon, and target it to the cell membrane. We performed transient transfection of a RIBEYE construct with a C-terminal EGFP tag in HEK293 cells and observed large spherical clusters of RIBEYE that appeared largely cytosolic (Fig. 1Aii), in contrast to a diffuse cytosolic distribution when merely expressing EGFP (Fig. 1Ai). These RIBEYE clusters form due to self-assembling properties of RIBEYE via multiple sites of homophilic interaction as have been demonstrated before in several cell lines. They do not colocalize with the endoplasmic reticulum (ER), Golgi apparatus, or lysosomes and hence, are unlikely to reflect RIBEYE entrapped in protein trafficking pathways or to represent degradation products of overexpressed RIBEYE (Supplementary Fig. 1A, B and C).

Membrane targeting of Bassoon using a palmitoylation consensus sequence.

A Representative confocal images of HEK293 cells transfected with (i) GFP only (CtBP2/RIBEYE in magenta labeling nuclei, GFP in cyan), (ii) RIBEYE-GFP (CtBP2/RIBEYE in magenta, GFP in cyan), (iii) Bassoon (green) and (iv) RIBEYE-GFP and Bassoon. Note that antibodies against RIBEYE B-domain and the nuclear transcription factor CtBP2 result in similar staining patterns as CtBP2 is transcribed from the same gene as RIBEYE and is identical to the RIBEYE-B domain except for the first 20 N-terminal amino acids. Upper panel shows maximum projections and lower panel shows exemplary single sections from confocal stacks. Scale bar = 5 µm.

B Schematic of construct for membrane targeting of Bassoon. The first 95 amino acids from full-length Bassoon were replaced with a palmitoylation consensus sequence from GAP43. Constructs without and with a C-terminal GFP tag were used, as depicted.

C Sample confocal image (single section) showing membrane targeted palm-Bassoon (green) expressed in HEK293 cells appears as puncta distributed along the periphery of the cell, marked by Na, K-ATPase α1(blue). Inset shows maximal projection of confocal section. Scale bar = 5µm.

D Zoom-in from (C) shows colocalisation of palm-Bassoon with membrane marker Na, K-ATPase α1. Scale bar = 1µm.

E Quantification of Bassoon signal intensity at periphery vs inside of cell. Cells expressing palm-Bassoon (N = 10 cells) clearly show a higher peripheral distribution compared to cells expressing full-length Bassoon (N = 10 cells), ****P < 0.0001, Mann-Whitney-Wilcoxon test. Overlaid data points represent individual cells, crosses represent the mean values, central band indicates the median, whiskers represent 90/10 percentiles, and boxes represent 75/25 percentiles.

Next, for membrane targeting of these cytosolic RIBEYE clusters, we co-expressed the multidomain cytomatrix of the active zone protein Bassoon (Fig. 1Aiii). Prior work on the molecular underpinnings of ribbon synapses had identified Bassoon (tom Dieck et al, 1998) to critically contribute to anchoring the synaptic ribbon to the AZ membrane (Khimich et al, 2005; Dick et al, 2003; tom Dieck et al, 2005). Co-expression of full-length Bassoon along with RIBEYE in HEK293 cells showed colocalizing clusters of the two proteins that, however, remained largely cytosolic (Fig. 1Aiv).

Next, for plasma membrane-targeting of Bassoon, we generated a construct by removing the first 95 N-terminal amino acids of Bassoon and replacing these with a palmitoylation consensus sequence of the neuronal Growth Associated Protein 43 (GAP43). We refer to this as ‘palm-Bassoon’ throughout, and we used constructs with and without a C-terminal EGFP tag (Fig. 1B). Expression of either of these palm-Bassoon constructs in HEK293 cells showed comparable immunofluorescence patterns with Bassoon puncta spread across the periphery of the cell, largely colocalizing with the endogenously expressed plasma membrane-standing Na, K-ATPase α1 (data representative of 3 transfections, Fig. 1C, D). Palm-Bassoon does not appear to localize in the ER, Golgi apparatus or lysosomes (Supplementary Fig. 1D, E and F). Comparing the ratio of Bassoon signal intensity at the periphery versus inside of the cell in randomly selected single sections from confocal stacks of palm-Bassoon and Bassoon transfected cells (N = 10 cells, 3 transfections per group, Fig. 1E) demonstrated a higher Bassoon signal intensity at the periphery of cells expressing the palm-Bassoon construct (****P < 0.0001, Mann-Whitney-Wilcoxon test), implying successful plasma membrane targeting of Bassoon.

Next, we co-expressed palm-Bassoon and RIBEYE in HEK293 cells and observed colocalizing RIBEYE and Bassoon immunofluorescent puncta at the periphery of cells (7 transfections; Fig. 2Ai). Closer inspection of these immunofluorescent puncta with STimulated Emission Depletion (STED) nanoscopy (Fig. 2Aii) revealed discrete structures typically consisting of ellipsoidal RIBEYE clusters juxtaposing on top of plate-like palm-Bassoon structures which seemingly anchor the RIBEYE clusters to the plasma membrane. We found the morphology of the RIBEYE + palm-Bassoon structures to be strikingly reminiscent of the arrangement of the two proteins in IHC ribbon synapses, where an ellipsoid/spherical synaptic ribbon composed of RIBEYE is found seated on a Bassoon plate that anchors it to the presynaptic AZ membrane (e.g. Wong et al, 2014; Michanski et al, 2019, 2023) (Fig. 2B, i and ii; data from Michanski et al, 2023).

Co-expression of RIBEYE with palm-Bassoon results in ribbon-type AZ-like structures

A (i) Representative confocal image (single section) of a HEK293 cell transfected with RIBEYE-GFP (magenta) and palm-Bassoon (green). Co-expression of RIBEYE with palm-Bassoon targets RIBEYE to the cell membrane. Inset shows maximum projection. Scale bar = 5µm. (ii) Exemplary 2-D STED images for RIBEYE – palm-Bassoon juxtapositions acquired from cells as shown in (i); Scale bar = 500nm; individual channels have been depicted with an intensity-coded look-up table with warmer colors indicating higher intensity.

B (i) Representative maximum projection of confocal sections of apical organ of Corti from a Wistar rat (postnatal day 18); data as published in Michanski et al, 2023, stainings were for CtBP2/RIBEYE (magenta; labeling synaptic ribbons and IHC nuclei) and Bassoon (green; spots juxtaposing with ribbons represent IHC AZs, spots not juxtaposing with ribbons represent efferent synapses formed by lateral olivocochlear neurons onto SGN boutons). Scale bar = 5µm. (ii) Juxtaposing RIBEYE and Bassoon spots imaged in 2-D STED and confocal mode respectively. Scale bar = 500nm. Note the striking resemblance to reconstituted RIBEYE + palm-Bassoon structures in HEK293 cells as shown in (Aii). STED images are from 3 sample transfections, representative of 7 total transfections; individual channels have been depicted with an intensity-coded look-up table with warmer colors indicating higher intensity.

C Quantification of RIBEYE and Bassoon signal intensity at periphery vs inside of cell shows a higher peripheral distribution of RIBEYE and Bassoon in HEK293 + RIBEYE + palm-Bassoon cells (N=9 cells) as compared to HEK293 + RIBEYE + Bassoon cells (N=9 cells); ****P < 0.0001, Mann-Whitney-Wilcoxon test. Overlaid data points represent individual cells, crosses represent mean values, central band indicates the median, whiskers represent 90/10 percentiles, and boxes represent 75/25 percentiles.

D Distribution of volumes of RIBEYE and palm-Bassoon puncta from HEK293 cells expressing RIBEYE and palm-Bassoon (n = 20 cells, quantifications from 5 sample transfections). Volumes of synaptic ribbons from rat IHCs (n = 29 cells, 3 rats) have been plotted for comparison.

E Box plot depicting data from (D). Volumes of RIBEYE and palm-Bassoon puncta are comparable to each other (P > 0.99, Kruskal-Wallis test with post hoc Dunn’s multi-comparison), but on an average much smaller and considerably more variable when compared to volumes of RIBEYE and Bassoon puncta from rat IHCs respectively (****P < 0.0001, Kruskal-Wallis test with post hoc Dunn’s multi-comparison). Overlaid plus signs represent individual spots, crosses represent mean values, central band indicates the median, whiskers represent 90/10 percentiles and boxes represent 75/25 percentiles.

F RIBEYE puncta in HEK cells expressing RIBEYE and palm-Bassoon (nspots = 864, N = 20 cells) appear more spherical than Bassoon puncta in the same cells (nspots = 961, N = 20 cells; ****P < 0.0001, Kruskal-Wallis test with post hoc Dunn’s multiple comparison test). Note the similar trend in IHC synaptic ribbons where RIBEYE puncta are more spherical than Bassoon puncta (nspots = 290 for RIBEYE and nspots = 314 for Bassoon, N = 29 cells, 3 rats; ****P < 0.0001, Kruskal-Wallis test with post hoc Dunn’s multiple comparison test). Overlaid plus signs represent individual spots, crosses represent mean values, central band indicates the median, whiskers represent 90/10 percentiles and boxes represent 75/25 percentiles.

We compared the RIBEYE and Bassoon signal intensities at the periphery versus inside of the cell in randomly selected single sections from confocal stacks of HEK cells expressing RIBEYE and palm-Bassoon (N = 9 cells) and HEK cells expressing RIBEYE and full-length Bassoon (N=9 cells). We found an increased peripheral distribution of both RIBEYE and Bassoon when using palm-Bassoon (Fig. 2C, ****P < 0.0001, Mann-Whitney-Wilcoxon test), implying successful membrane targeting of RIBEYE by palm-Bassoon. In each transfection, ∼10% cells showed co-expression of both RIBEYE and palm-Bassoon (for representative sample overview see Supplementary Fig. 2). Of those, cells expressing discrete RIBEYE-palm Bassoon structures were discernible by the characteristic RIBEYE distribution along the periphery of the cell. This peripheral distribution in turn seemingly depends upon RIBEYE and palm-Bassoon expression ratios (shown in Supplementary Fig. 2B) as cells with little to no palm-Bassoon expression show predominantly cytosolic RIBEYE puncta and were not used for analysis. For simplicity, we henceforth refer to the structures composed of RIBEYE and palm-Bassoon in HEK293 cells as SyRibbons (for synthetic ribbons).

We next performed three-dimensional surface renderings using Imaris 9.6 (Oxford Instruments) to assess these structures. In a given cell, only structures with colocalizing RIBEYE and palm-Bassoon immunofluorescence were considered for analysis to exclude occasional non-membrane localized spots (on average 51.75 ± 40.84 RIBEYE surfaces colocalizing with Bassoon surfaces per cell; N = 20 cells). The volumes of RIBEYE and palm-Bassoon surfaces of SyRibbons were smaller on average and more variable (average volume ± standard deviation (SD) = 0.19 ± 0.23 µm3 with a coefficient of variation (CV) = 1.23 for RIBEYE; nspots = 864 and volume = 0.17 ± 0.18 µm3 with CV = 1.10 for Bassoon; nspots = 951; data from N = 20 cells, quantifications from 5 sample transfections) than volumes of synaptic ribbons and Bassoon immunofluorescent puncta from rat IHCs (volume = 0.29 ± 0.17 µm3, CV = 0.59 for RIBEYE; nspots = 290, and volume = 0.27 ± 0.18 µm3, CV = 0.67 for Bassoon; nspots = 314, data from N = 29 cells), (Fig. 2D, E). Moreover, volumes of RIBEYE surfaces show a high positive correlation to volumes of corresponding palm-Bassoon surfaces (Pr = 0.778, ****P < 0.0001), implying that larger palm-Bassoon structures may recruit bigger RIBEYE structures to the plasma membrane. We also note that the volume of RIBEYE surfaces in SyRibbons appears smaller and well-regulated in contrast to the predominantly large, cytosolic RIBEYE assemblies in cells co-expressing RIBEYE and full-length Bassoon (volume = 0.30 ± 0.36 µm3, nspots = 460, N = 10 cells, ****P < 0.001, Mann-Whitney-Wilcoxon Test). In turn, Bassoon clusters seemed regulated by co-expressed RIBEYE: Bassoon surfaces at the plasma membrane were larger at SyRibbons than in the absence of RIBEYE in cells only expressing palm-Bassoon (volume = 0.14 ± 0.21 µm3, nspots = 2217, N = 13 cells, ****P < 0.001, Mann-Whitney-Wilcoxon Test). RIBEYE clusters constituting SyRibbons appeared more spherical than the plate-like Bassoon structures in the same cells (sphericity Ψ = 0.91 ± 0.05, nspots = 864 for RIBEYE versus Ψ = 0.86 ± 0.07, nspots= 951 for Bassoon, data from N = 20 cells; ****P < 0.0001, Kruskal-Wallis test with post hoc Dunn’s multiple comparison test). This follows the same trend as in rat IHCs where RIBEYE spots indeed appear more spherical (Ψ = 0.90 ± 0.05, nspots = 290) compared to Bassoon spots (Ψ = 0.76 ± 0.10, nspots = 314); ****P < 0.0001, Kruskal-Wallis test with post hoc Dunn’s multiple comparison test, data from N = 29 cells (Fig. 2F). Nonetheless, the fact that next to structures with volumes comparable to IHC ribbons, we also encountered smaller and larger structures, likely reflects poorer regulation of RIBEYE and palm-Bassoon expression in the heterologous system.

Next, we analyzed SyRibbons in situ using cryo-electron tomography (cryo-ET), which capitalizes on cell vitrification to obtain near-native preservation. Plunge-frozen HEK cells transfected with RIBEYE-GFP and untagged palm-Bassoon were vitrified and subsequently milled using a cryo-focused ion beam (cryo-FIB; Supplementary Fig. 3A) to produce 150 nm-thick lamellae as previously described (Rigort et al, 2010; Pierson et al, 2024, see methods). Using a light microscope integrated in the cryo-FIB chamber, we targeted cell areas showing peripheral GFP fluorescence corresponding to SyRibbons (Supplementary Fig. 3B). Lamellae were subsequently transferred to a cryo-transmission electron microscope, and tomographic tilt-series were acquired at fluorescence locations. The tomograms showed that GFP-positive RIBEYE spots corresponded to electron-dense structures (Fig 3A-B), as is characteristic of the synaptic ribbon (Robertis & Franchi, 1956; Smith & Sjöstrand, 1961). These electron-dense SyRibbons appeared to be approx. 300 – 800 nm in size and were ovoid or ellipsoidal in shape (Fig 3B-C, Supplementary Fig. 3C). Some SyRibbons displayed a hollow core (Supplementary Fig. 3Ci and 3Ciii) and on closer inspection, some appeared to have a multi-lamellar ultrastructure (Supplementary Fig. 3Cii), both of which have been reported previously for natively expressed synaptic ribbons in the inner ear and the retina (Michanski et al, 2023, 2019; Sobkowicz et al, 1982; Liberman, 1980; Stamataki et al, 2006; Wichmann & Moser, 2015). We captured a SyRibbon positioned within 100 nm from the plasma membrane (Fig 3C). Although bona fide ribbons display a halo of SVs tethered on their surface, we did not observe any obvious accumulation of vesicular structures around these SyRibbons, which is not unexpected for synapse-naïve HEK cells. This is in contrast to a previous report in R28 retinal progenitor cells, where heterologously expressed RIBEYE was shown to recruit vesicles (Magupalli et al, 2008). Intriguingly, in some tomograms, we observed membrane-bound SyRibbons (Supplementary Fig. 3Ci-iv), which may indicate an autophagic engulfment of some of these overexpressed structures.

Cryo-correlative microscopy captures membrane-localised SyRibbons.

A. RIBEYE-GFP signal on a 150 nm thick lamella, revealed by fluorescent light microscopy within the cryo-FIB chamber (cryo-LM). Dotted lines delineate cell membranes. The orange arrow indicates a plasma membrane-proximal GFP fluorescence, whereas purple arrow heads point to cytosolic RIBEYE aggregates. The tomogram shown in (C) was acquired at the boxed region. The image underwent background subtraction for better visualization.

B. Transmission electron microscopy (TEM) image of the lamella in A. Orange arrow and purple arrowheads locate the respective regions in (A), showing that GFP-positive spots correlated to electron dense bodies. Dotted lines delineate plasma membranes. The tomogram shown in (C) was taken at the boxed region.

C. Tomogram slice showing a SyRibbon acquired at the boxed region in (B). PM: plasma membrane, ECS: extracellular space ER: endoplasmic reticulum.

RBP2 tethers CaV1.3 channels

To study the potential of SyRibbons to cluster Ca2+ channels, we employed HEK293 cells stably expressing an inducible transgene of CaV1.3α1 along with constitutive transgenes for CaVβ3 and CaVα281. We either used tetracycline for inducible CaV1.3α1 channel expression (Fig. 4A) or performed transient transfections with CaV1.3 constructs containing an N-terminal EGFP- (Fig. 4B) or Halo-tag (Fig. 4C) for direct fluorescence imaging. Expression of either Ca2+ channel complex resulted in small clusters at the plasma membrane as represented in Fig. 4D (average volume 0.08 ± 0.13 µm3, nspots = 949, N = 12 cells for untagged inducible CaV1.3α1; 0.07 ± 0.08 µm3, nspots = 392, N = 11 cells for EGFP-CaV1.3; and 0.07 ± 0.09 µm3, nspots = 1281, N = 16 cells for Halo-CaV1.3α; Puntagged/EGFP = 0.547, Puntagged/Halo > 0.999, PHalo/EGFP = 0.582, Kruskal-Wallis test with post hoc Dunn’s multiple comparison test).

HEK293 cells expressing CaV1.3

A Representative confocal section of a HEK293 cell stably expressing an inducible CaV1.3α transgene (untagged) along with constitutive transgenes for CaVβ3 and CaVα281. Immunostainings have been performed using antibodies against CaV1.3 (green) and Na, K-ATPase α1 (blue, labeling the plasma membrane). Scale bar = 10 µm.

B Representative confocal section of a HEK293 cell stably expressing CaVβ3 and CaVα281, and transiently transfected with an N-terminal EGFP-tagged CaV1.3 construct. Immunostainings have been performed using antibodies against GFP (green) and Na, K-ATPase α1 (blue). Scale bar = 10 µm.

C Representative confocal section of a HEK293 cell stably expressing CaVβ3 and CaVα281 and transiently transfected with an N-terminal Halo-tagged CaV1.3 construct. Immunostainings have been performed using antibodies against CaV1.3 (green) and Halo tag (magenta). Scale bar = 10 µm.

D Expression of either of the Ca2+ channel complexes result in clusters of comparable sizes, (P > 0.05, Kruskal-Wallis test with post hoc Dunn’s multiple comparison test). Overlaid plus signs represent individual spots, crosses represent mean values, central band indicates the median, whiskers represent 90/10 percentiles and boxes represent 75/25 percentiles.

Co-expression of palm-Bassoon did not show colocalization with CaV1.3, which is in line with observations with full-length Bassoon (Frank et al, 2010) and the same applied when co-expressing RIBEYE and CaV1.3, arguing against a direct interaction for both (Supplementary Fig. 4). Rab-interacting molecule-binding protein (RIM-BP or RBP) links Bassoon to Ca2+ channels (Davydova et al, 2014) and hence was an interesting candidate for tethering CaV1.3 to Bassoon clusters. RBP has been shown to interact with CaV1.3 (Hibino et al, 2002) and to be required for normal CaV1.3 Ca2+ channel clustering at the IHC ribbon synapse (Krinner et al, 2017, 2021). Indeed, when co-expressing RBP2 with palm-Bassoon and CaV1.3, we found juxtaposed palm-Bassoon and CaV1.3 both with immunolabeled (Fig. 5A) and live-labeled (Fig. 5B, C) proteins indicating successful clustering of CaV1.3 at synthetic AZs.

RBP2 bridges palm-Bassoon and CaV1.3 to form supra-molecular assemblies at the plasma membrane.

A Representative confocal images (single sections) of HEK293 cells (fixed) transfected with palm-Bassoon (green), RBP2 (red) and N-terminal Halo-tagged CaV1.3 (blue). The three colocalizing proteins appear to form supra-molecular assemblies at the cell membrane. Scale bar = 5 µm for left panel showing cell overviews, 2 µm for right panels showing zoom-ins.

B Representative confocal images (single sections) of live HEK293 cells transfected with palm-Bassoon-GFP (green), RBP2-p2A-mKATE2 (red, representing only mKATE2 signal and not RBP2 localization) and Halo-CaV1.3 (blue, labeled by with JF-646 HaloTag Ligand before imaging). Scale bar = 5 µm.

C Exemplary super-resolution STED images from live-labeled samples in (B) showing juxtaposition of palm-Bassoon and Halo-CaV1.3 signal at the plasma membrane in RBP2 positive cells. Scale bar = 1 µm.

We next performed ruptured whole-cell patch-clamp to assess if RBP2 and/or palm-Bassoon or RIBEYE co-expression results in changes in CaV1.3 Ca2+ currents. After tetracycline treatment for CaV1.3α1 expression (18-24 hours), cells were transfected with both RBP2-p2A-mKATE2 and palm-Bassoon-GFP, with only palm-Bassoon-GFP, only RBP2-p2A-mKATE2 or RIBEYE-GFP constructs (Supplementary Fig. 5A). After 18-24 hours, the cell culture media was changed, and cells were recorded 24-36 hours afterwards. We used 10 mM [Ca2+]e and recorded Ca2+ current (density) - voltage relations (IVs) ∼1 minute after establishing the whole-cell configuration by applying step depolarizations of 20 ms from −86.2 mV to 58.8 mV in 5 mV increments. Ca2+ current density appeared to be augmented upon expression of both RBP2 and palm-Bassoon when compared with induced-only controls (Supplementary Fig. 5B, D; **P = 0.0041, Kruskal-Wallis test with post hoc Dunn’s multiple comparison test). However, Ca2+ current density in cells expressing only RBP2 or Palm-Bassoon did not significantly differ from either induced-only controls or cells expressing both proteins (P > 0.05, Kruskal-Wallis test with post hoc Dunn’s multiple comparison test). For cells expressing only RIBEYE, we also did not observe any changes (P > 0.999, Kruskal-Wallis test with post hoc Dunn’s multiple comparison test). We did not observe any noticeable differences in voltage-dependence, activation kinetics or inactivation kinetics of CaV1.3 Ca2+ current upon expression of any of these AZ proteins (Supplementary Fig. 5C, E, F; Supplementary Fig. 6A, B).

Larger synthetic ribbon-type AZs establish larger CaV1.3 Ca2+ channel clusters

Next, we tetra-transfected HEK293 cells with palm-Bassoon, Halo-tagged CaV1.3α1, RBP2 and RIBEYE to explore the impact of co-expressing of RIBEYE on the clustering of CaV1.3 Ca2+ channels (Fig. 6A for transfection scheme). Using confocal imaging of immunofluorescently labelled RIBEYE, Halo-CaV1.3 and Bassoon (RBP2 immunofluorescence was imaged using epifluorescence to select tetra-transfected HEK293 cells), we observed colocalization of CaV1.3α1, Bassoon and RIBEYE (Fig. 6B). Line profiles drawn tangentially along the plasma membrane showed that beyond the general distribution of a CaV1.3α1 signal in the plasma membrane, CaV1.3α1 signal hotspots occurred underneath the SyRibbons (Fig. 6C). The signal intensity of RIBEYE immunofluorescence positively correlated with the CaV1.3 signal immunofluorescence, (Pr = 0.839 and 0.886 from two sample line scans, ****P < 0.0001), which has previously been reported for immunolabeled native ribbon-type AZs of mouse IHCs (Ohn et al, 2016). We further performed a pixel-based correlation analysis (Supplementary Fig. 7A, B) which indicated a moderate to strong colocalisation of RIBEYE and CaV1.3 upon co-expression of palm-Bassoon and RBP2. In contrast, a poor colocalization was found in cells expressing only RIBEYE and CaV1.3; average Pearson’s colocalisation coefficient = 0.41 versus 0.24; average Mander’s overlap coefficient = 0.63 versus 0.34, N = 12 and 8 cells respectively (**P < 0.01m Wilcoxon’s Rank Test). Colocalisation quantifications for other transfection combinations shown in Fig. 2, 5 and Supplementary Fig. 4 have been collectively depicted in Supplementary Fig. 7. We next turned to 2-color STED imaging which provided improved spatial resolution and highlighted the confined localization of CaV1.3 Ca2+ -channel clusters underneath SyRibbons as shown in Fig. 6D.

Synthetic ribbon-type active zones recruit CaV1.3 Ca2+ channels.

A Experimental scheme for expression of Halo tagged CaV1.3, palm-Bassoon, RIBEYE-GFP and RBP2 in HEK293 cells.

B Representative confocal images (single sections) of HEK293 cells transfected with RBP2 (not shown, expression confirmed using epifluorescence), RIBEYE-GFP (red), palm-Bassoon (green), Halo-CaV1.3 (blue) shows colocalization of the three latter proteins at the plasma membrane. Scale bar = 5 µm.

C Confocal zoom-ins from (B). Line scans depict an increased CaV1.3 signal intensity (blue) at sites where RIBEYE (red) and palm-Bassoon (green) clusters localize. Pearson’s correlation coefficients (Pr) were calculated along the line profiles for RIBEYE and CaV1.3 signal intensity and indicate a high degree of correlation (∼0.8) between the spatial localization of the two. Scale bar = 2 µm.

D Super-resolution STED images showing CaV1.3 clusters localizing at the base of the SyRibbons. Scale bar = 500nm.

E Distribution of volumes from confocal images of CaV1.3 clusters colocalizing with SyRibbons (orange, nspots = 634) and not colocalizing with SyRibbons (yellow, nspots = 1354) from HEK293 cells expressing CaV1.3, RBP2 and SyRibbons (N = 13 cells). Pooled volumes of all CaV1.3 clusters (with and without SyRibbons) from these cells have been shown in red (nspots = 1988). Volumes of CaV1.3 clusters from HEK293 cells expressing only Halo-CaV1.3 (black, nspots = 1281, N = 16 cells) and from cells expressing Halo-CaV1.3, RBP2 and palm-Bassoon (blue, nspots = 427, N = 6 cells) have been plotted for comparison.

F Box plot depicting data from (E). CaV1.3 clusters appear larger on an average in cells expressing CaV1.3, RBP2 and SyRibbons versus in cells expressing only CaV1.3, (**P = 0.0044). Within cells expressing all four proteins, CaV1.3 clusters are much larger when colocalizing with SyRibbons versus when they do not (****P < 0.0001). CaV1.3 clusters colocalising with SyRibbons were in fact also larger than clusters in cells expressing only CaV1.3 and cells expressing only RBP2 and palm-Bassoon (****P < 0.0001). Statistical test: Kruskal-Wallis test with post hoc Dunn’s multiple comparison correction. Overlaid plus signs represent individual spots, crosses represent mean values, central band indicates the median, whiskers represent 90/10 percentiles and boxes represent 75/25 percentiles.

We performed surface renderings of confocally imaged Halo-tagged CaV1.3α1, RIBEYE and palm-Bassoon immunofluorescence spots as described above for estimating the volume of SyRibbons and the CaV1.3 Ca2+ channel clusters (Fig. 6E, F). This revealed significantly larger CaV1.3 Ca2+ channel clusters when colocalizing with SyRibbons in HEK293 + Halo-CaV1.3 + RBP2 + SyRibbons cells (average volume ± SD = 0.18 ± 0.28 µm3, nspots= 634, N = 13 cells, data representative of 5 transfections) compared to ribbonless (i.e. non-colocalized) CaV1.3 Ca2+ channel clusters from the same cells (volume = 0.06 ± 0.10 µm3, nspots= 1354, N = 13 cells; ****P < 0.0001, Kruskal-Wallis test with post hoc Dunn’s multiple comparison test) and CaV1.3 Ca2+ clusters in HEK293 cells solely expressing Halo-CaV1.3 Ca2+ channels (volume = 0.07 ± 0.09µm3, nspots= 1281, N = 16 cells, data representative of 4 transfections; ****P < 0.0001, Kruskal-Wallis test with post hoc Dunn’s multiple comparison test). Ribbonless CaV1.3 Ca2+ channel clusters did not differ significantly regardless if co-expressing RIBEYE (P = 0.099, Kruskal-Wallis test with post hoc Dunn’s multiple comparison test, see Fig. 6F). Moreover, volumes of CaV1.3 cluster surfaces show a moderate positive correlation with volumes of corresponding RIBEYE surfaces (Pr = 0.385, ****P < 0.0001), similar to previous observations at IHC ribbon synapses (Frank et al, 2009; Michanski et al, 2023; Ohn et al, 2016).

Ca2+ imaging reveals a partial spatial confinement of Ca2+ signal at synthetic ribbon-type active zones

Finally, we set-out to functionally characterize synthetic ribbon-type AZs. We combined patch-clamp recordings of CaV1.3 Ca2+ influx with spinning-disk confocal microscopy to visualize Ca2+ signals at the synthetic ribbon-type AZs in HEK293 cells. Following tetracycline-induction for CaV1.3 expression, HEK293 cells were co-transfected with constructs expressing RIBEYE-GFP, untagged palm-Bassoon and RBP2-p2A-mKATE2, (Fig. 7A). We recorded mKATE2 positive cells that expressed peripheral GFP puncta, indicative of RIBEYE and palm-Bassoon co-expression. We employed peripheral GFP expression as a proxy of SyRibbons for functional analysis (Fig. 7B). Ruptured patch-clamp recordings were performed with 10 mM [Ca2+]e and 10 mM intracellular EGTA to enhance the signal to background ratio for visualizing Ca2+ influx using the low-affinity red-shifted Ca2+ indicator Calbryte590 (100 µM, kd = 1.4 µM). After loading the cell for approximately a minute, we first applied step depolarizations of 20 ms from −83 to +62 mV in 5-mV step increments to measure Ca2+ current (density)–voltage relations. We did not observe any changes in either the Ca2+ current density (P = 0.293, t-test) or the voltage-dependence of Ca2+ current (P = 0.780, t-test when comparing Vhalf) when comparing cells expressing CaV1.3 + RBP2 + SyRibbons with cells expressing only CaV1.3 (Fig. 7C and 7D). Next, we applied a depolarizing pulse to +2 mV for 500 ms and acquired images at a frame rate of 20 Hz (Fig. 7E). The intracellular Ca2+ signals mediated by CaV1.3 Ca2+ influx were visualized as an increase in Calbryte590 fluorescence. The increase of Calbryte590 fluorescence appeared to spread throughout the HEK293 cell membrane with discrete regions of higher signal intensity at SyRibbons (exemplary cell shown in Fig. 7F with zoom-in at site with SyRibbon shown in 7G). We note that we did not find noticeable red fluorescence at SyRibbons in the absence of depolarization, which is similar to findings with Fluo-4FF in IHCs (Frank et al, 2009), but different from the situation for the large spherical presynaptic bodies in bullfrog hair cells that are stained by Fluo-3 (Issa & Hudspeth, 1996). Line profiles drawn tangentially to the membrane in composite ΔF image of Calbryte590 signal and RIBEYE-GFP showed a high correlation between the localization of SyRibbons and peak intensity of Calbryte590 fluorescence increase, indicating preferential Ca2+ signaling at the SyRibbons (two representative line scans shown in Fig. 7H, Pearson correlation coefficients of 0.783 and 0.757). Recordings were made from 24 cells (7 transfections) and ROIs (diameter = 2 µm) with and without SyRibbons were analyzed as shown in Fig. 7I. Calbryte590 ΔFmax/F0 values were calculated for each ROI which on an average showed higher ΔFmax/F0 values for ROIs with SyRibbons (on average ∼26% higher) than for ROIs without SyRibbons (*P = 0.018, paired t-test, Fig. 7J). We note that occasionally in some cells we observed a higher Calbryte590 signal increase at sites without SyRibbons, and we attribute this to the likely presence of AZ-like clusters composed of RBP2 and palm-Bassoon in these regions which we could not visualize.

Ca2+ imaging reveals higher Ca2+ signal intensity underneath SyRibbons.

A Experimental scheme for expression of SyRibbons, RBP2 and CaV1.3 in HEK293 cells for patch clamp in combination with Ca2+ imaging using the low affinity Ca2+ indicator Calbryte590.

B Exemplary cell used for Ca2+ imaging. Cells were identified by peripheral RIBEYE-GFP puncta as shown (indicative of RIBEYE and palm-Bassoon co-expression) and mKATE2 signal (indicative of RBP2 expression).

C Current density-voltage (IV) relations from whole-cell patch clamp recordings of HEK293 cells expressing CaV1.3 + RBP2 + SyRibbons (n=20, in orange) and only CaV1.3 (n=10, in black). Lines represent mean current traces, shaded area represents ± SEM. Whole-cell Ca2+ current density does not appear to change upon co-expression of RBP2 and SyRibbons (P = 0.293, t-test).

D The voltage-dependence of Ca2+ current influx does not seem to be altered upon co-expression of RBP2 and SyRibbons at the whole-cell level.

E A depolarizing pulse to +2mV was applied to the cells for 500 ms and the increase in Calbryte590 fluorescence was measured by acquiring images at a frame rate of 20 Hz using a spinning disk confocal microscope.

F Representative image of a HEK293 cell expressing SyRibbons, RBP2 and CaV1.3 stimulated as described in (E). RIBEYE-GFP signal has been shown in magenta; the change in Calbryte590 fluorescence intensity (ΔF) upon Ca2+ binding (the frame at the onset of depolarization to +2mV) has been shown in green. Note the distinct Ca2+ signal “hotspots” that co-localize with SyRibbons. Scale bar = 10 µm.

G Zoom-ins from (F). Frames before, at the onset of, and after stimulation have been shown. Note the localized Ca2+ influx at the base of the SyRibbons. Scale bar = 1 µm.

H Line profiles drawn tangentially to the membrane in composite ΔF image of Calbryte590 signal and RIBEYE-GFP show a high correlation between the localization of SyRibbons and peak intensity of Calbryte590 fluorescence increase (Pr = ∼0.7). Plots show intensity profiles along line scans for RIBEYE-GFP (magenta) and Calbryte590 (light grey for frames before stimulus, bold green for peak intensity at onset of stimulus, light green lines for decaying intensity during ongoing stimulus and dark grey for frames post-stimulus).

I Regions of interest (ROIs) of 2 µm diameter were drawn at sites with (magenta circles) and without (green circles) SyRibbons and the corresponding Calbryte590 ΔF/F0 values were calculated for each ROI.

J Plot of average (shaded area represents ± SEM) ΔF/F0 values from ROIs with and without SyRibbons (218 and 139 ROIs respectively from N = 24 cells, data from 7 transfections).

K On average, Calbryte590 ΔFmax/F0 was higher for ROIs with SyRibbons than ROIs without them in a given cell (*P = 0.018, paired t-test). Dots represent mean of ΔFmax/F0 from ROIs with (magenta) and without (green) SyRibbons averaged per cell, error bars represent ± SEM.

Discussion

In this study, we reconstituted and characterized a minimal ribbon-type AZ model system using heterologous expression. Co-expressing CaV1.3 Ca2+ channels with membrane targeted Bassoon for RIBEYE anchorage to the plasma membrane and attracting exogenous RBP2 for clustering Ca2+ channels in HEK293 cells led to structures recapitulating basic aspects of IHC ribbon synapses. Despite the poor regulation of transgene expression in the synapse-naïve HEK293 cells, the AZ-like structures, in the subset of cells were similar in morphology to the native IHC AZs. This applied to the synthetic ribbons as well as to the AZ-like clusters of Bassoon, RBP2 and CaV1.3 Ca2+ channels. However, all three components exhibited variability beyond the substantial natural heterogeneity found among the IHC AZs (Moser et al, 2023). Although expression of RIBEYE does not seem to affect CaV1.3 physiology and kinetics, we demonstrate enhanced clustering of CaV1.3 channels and localization of Ca2+ influx at SyRibbons, with a mild increase in Ca2+ signal intensity. As SyRibbons partially resemble native hair cell AZs, we expect this easily available and experimentally accessible system to serve as an advanced testbed for functional interactions of AZ proteins and Ca2+ channels. Further efforts towards reconstituting SVs and their release sites will help to enhance the utility of these synthetic AZs.

Synthetic active zone reconstitution model to study ribbon synapse assembly

By demonstrating that membrane-targeted Bassoon suffices to anchor ribbon-like RIBEYE assemblies, this work adds to the top-down and bottom-up evidence for a key role of Bassoon in attracting RIBEYE to the presynaptic density at the AZ in photoreceptors and hair cells (Dick et al, 2003; Khimich et al, 2005; Jing et al, 2013; tom Dieck et al, 2005). Additional expression of multi-domain proteins of the presynaptic density such as CAST (Ohtsuka et al, 2002; Inoue et al, 2006) might avoid the need for artificial palmitoylation of Bassoon for membrane targeting. Furthermore, our results support a model of a bidirectional control of AZ size and shape by RIBEYE and proteins of the presynaptic density. The amount of RIBEYE assembled in membrane localized SyRibbons appeared regulated strongly contrasting the seemingly unregulated cytosolic RIBEYE assemblies dominating HEK293 cells in the absence of membrane-targeted Bassoon. Vice versa, clusters of Bassoon at the plasma membrane were larger at SyRibbons than in the absence of RIBEYE. Finally, the size of SyRibbons seemingly scaled with the size of CaV1.3 and Bassoon clusters similar to what is observed in IHCs (Ohn et al, 2016). Such a model is consistent with results from genetic perturbation of native ribbon synapses: i) disruption of RIBEYE leads to a disintegration of Bassoon into smaller subclusters in IHCs (Jean et al, 2018), and ii) disruption of CAST and ELKS reduces the size of the ribbon-type AZs in rod photoreceptors (tom Dieck et al, 2012; Hagiwara et al, 2018). We speculate that achieving the full extent of large ribbon-type AZs such as in rod photoreceptors as well as regulating the size of ribbon-type AZs at a specific synapse according to the precise functional demands requires such a functional interplay between proteins of the presynaptic density and RIBEYE. We note that other ribbon-resident proteins such as piccolino likely contribute to this fine-tuning of the size and shape of ribbon-type AZs, which will be an exciting question for future studies. Indeed, disruption of piccolino in IHCs reduced the average ribbon size (Müller et al, 2019; Michanski et al, 2023). One exciting example of ribbon synapse heterogeneity manifests itself in IHCs where synapses show a spatial size gradient that likely relates to the diverse molecular and functional properties of the postsynaptic SGNs (reviewed in Moser et al, 2023).

Insights into presynaptic Ca2+ channel clustering and function

Current discussion of the clustering of CaV Ca2+ channels offers two different models: low affinity protein-protein interactions (e.g. Hibino et al, 2002; Kaeser et al, 2011) via specific domains and liquid-liquid phase separation involving intrinsically disordered domains (e.g. Heck et al, 2019). Work on ribbon synapses has considered layers of organizing CaV Ca2+ channels at the AZ that seem more compatible with the former model: i) “micro-clustering” to which the synaptic ribbon contributes as a super-scaffold (Jean et al, 2018; Neef et al, 2018; Frank et al, 2010; Maxeiner et al, 2016) and ii) tethering and “nano-clustering” by RBPs, RIMs, CAST and ELKS (Liu et al, 2011; Grabner et al, 2015; Jung et al, 2015; Krinner et al, 2017; Luo et al, 2017; Hagiwara et al, 2018). Moreover, there is converging evidence for nanoscale coupling of few CaV Ca2+ channels to SV release sites (Brandt, 2005; Jarsky et al, 2010; Wong et al, 2014; Maxeiner et al, 2016; Özçete & Moser, 2021; Grabner & Moser, 2021; Jaime Tobón & Moser, 2023) for which a role of the ribbon has been more controversial (e.g. Maxeiner et al, 2016; Grabner & Moser, 2021). MINFLUX nanoscopy of rod photoreceptor AZs recently revealed a well-ordered double-line array topography of CaV Ca2+ channels, RIM, Bassoon and ubMunc13-2 at the presynaptic membrane on both sides of the ribbon (Grabner et al, 2022).

Clearly, we could not fully reconstitute this complex organization with the minimal set of molecular players at SyRibbons in HEK293 cells. Yet, we demonstrate a positive effect of SyRibbons on local Ca2+ signaling, recapitulating microclustering of CaV1.3 Ca2+ channels at IHC ribbon synapses. However, we note that the functional impact of ribbon-mediated clustering of CaV1.3 Ca2+ channels remains unclear at the moment. Our Ca2+ imaging data seems more reminiscent of previous work on ribbon-type AZs of immature IHCs (Wong et al, 2014) where a broader distribution of CaV1.3 and ribbons go along with an immature state of the IHC AZs. We also note that the difference in Ca2+ signalling near or apart SyRibbons is rather subtle. We propose two explanations in this regard: (i) the use of an overexpression system diminishes the contrast of the Ca2+ signal between AZ-like domains and the regular membrane. Overexpressed CaV1.3 is likely distributed diffusely across the plasma membrane in these cells along with more localised CaV1.3 clusters observed at SyRibbons (see Fig. 6B lower panel). (ii) While the presence of SyRibbons appears to promote the formation of larger CaV1.3 channels and partially localise the Ca2+ signal, it is unlikely that SyRibbons recruit a larger Ca2+ channel complement. IHCs of RIBEYE KO mice were reported to show normal Ca2+ current amplitudes at the whole-cell and single-AZ level (Jean et al, 2018), while the spatial spread of the Ca2+ signal was more diffuse in the absence of the ribbon.

We propose that SyRibbons scaffold CaV1.3 channel nano-clusters in a non-native heterologous expression system resulting in an increased channel density in regions with SyRibbons. In support of this hypothesis, the Ca2+ signal averaged at SyRibbons showed a mildly increased maximal amplitude when compared with diffuse Ca2+ channel distribution in the absence of SyRibbons. Instead, the total number of channels may be unaffected, which would explain why Ca2+ current amplitudes at the whole-cell level are not impacted by expression of SyRibbons. A clear limitation in the study therefore arises from the use of an overexpression system, making it necessary to interpret results with caution and requiring in vivo and ex vivo validation. Future use of polycistronic gene expression or stable cell lines expressing all components could help with more consistent expression of the components and make the system become even more widely applicable.

Another limitation of the synthetic AZ system in HEK293 cells is the lack of SVs and the molecular machinery required for SV release. Nonetheless, our work paves the way for future studies including stable co-expression of AZ proteins including of the release site marker Munc13-1 (Sakamoto et al, 2018; Böhme et al, 2016) and potential delivery of SV machinery via co-expression (Park et al, 2021). Alternatively, using neurosecretory cells such as pheochromocytoma cells (PC12) or adrenal chromaffin cells in primary culture, offer the advantage of providing synaptic-like microvesicles and a large set of components of the presynaptic machinery such that expression of exogenous proteins might be limited to RIBEYE. On the other hand, the complex molecular background involving various types of CaV channels will likely complicate the interpretation. Regardless, SyRibbons in cultured cells offer great availability and as adherent cells, are well accessible to sophisticated analysis by techniques such MINFLUX nanoscopy and cryo-electron tomography.

Materials and Methods

HEK cell culture and transfections

In this study, we used human embryonic kidney (HEK293) cells stably expressing a tetracycline-inducible human CaV1.3 pore forming α1-D subunit transgene (CACNA1D, NM_000720.2), and showing constitutive expression of CaVβ3 (CACNB3, NM_000725.2) and CaVα28-1 (CACNA2D1, NM_000722.2). The stably expressing cell line was acquired from Charles River Laboratories, Cleveland, Ohio, USA (Cat. No. CT6232). The cells were cultured as per product protocols in Dulbecco’s Modified Eagle Medium (DMEM) containing GlutaMax, high glucose and pyruvate (Gibco, Life Tech., 31966047), supplemented with 10% Fetal Bovine Serum (FBS; Gibco, Life Tech., A5256701) and 100 units/ml of Penicillin-Streptomycin (Gibco, Life Tech., 15070063). Additionally, the media was supplemented with 0.6 µM of Isradipine (Sigma Aldrich, I6658) and the following selection antibiotics (in mg/ml): 0.005 Blasticidin (Invivogen, ant-bl-05), 0.50 Geneticin (G418 Sulfate; Gibco, Life Tech., 10131027), 0.10 Zeocin (InvivoGen, ant-zn-05) and 0.04 Hygromycin (Thermo Fisher, 10687010). Cells were cultured in a humidified incubator at 37°C with 5% (v/v) CO2 saturation. Cells were split at ∼70% confluency every 3-4 days to prevent adverse effects on cell growth and channel expression, by dissociating the cell using Accutase (Sigma-Aldrich, A6964). The passage number did not exceed more than 26 passages. For experiments, cells were grown in media lacking selection antibiotics. For induction of α1-D subunit expression, cells were treated with selection antibiotic-free medium containing 3 µg/ml tetracycline. For transfection, 6.8 µg polyethylenimine (PEI, 25kDa linear, Polysciences, 23966) was added along with a total of 2 µg of DNA to a final volume of 100 µl of DMEM (without FBS). The PEI/DNA mixture was thoroughly mixed and allowed to incubate for 30 min at room temperature before being added to the cells (30 µl/well in a 24-well-plate containing 500 µl media, cell confluency ∼50-70%). For all experiments with co-transfections, we used an equimolar ratio of DNA (total amount of DNA was kept 2 µg in a 100 µl transfection mix). After 18-24 hours, transfection media was replaced with fresh media devoid of selection antibiotics. Cells were used for experiments (immunocytochemistry, patch clamp and Ca2+ imaging) 24-48 hours after changing the media or as specified. If cell confluency was too high, cells were reseeded to an appropriate confluency and allowed to settle for at least 12 hours before commencement of experiments.

Expression vectors for RIBEYE, Bassoon, CaV1.3 and RBP2

The RIBEYE-GFP construct used consisted of a human RIBEYE cDNA cloned inframe into a pAAV-GFP vector driven by a hybrid CMV-enhancer-human-beta-actin promoter (CMV-HBA) and containing a downstream Woodchuck Hepatitis Virus (WHV) Posttranscriptional Regulatory Element (WPRE) for mRNA stabilization. All Bassoon constructs encode rat Bassoon. The untagged full-length constructs were cloned in the pCS2+ vector. For the generation of palm-Bassoon constructs, we used a pEGFP-C1 vector backbone with a CMV promoter and kanamycin resistance cassette. The insert comprised of a palmitoylation consensus sequence of GAP43 (ATGCTGTGCTGTATGAGAAGAACCAAACAGGTTGAAAAGAATGATGAGGACCAAAAGATTTCCGGACTCAGATCTCGAG), followed by a cDNA sequence encoding amino acids 95-3938 of rat Bassoon, and a C-terminal monomeric GFP. For the untagged version, the GFP tag was replaced by a STOP codon. To generate the Halo-CaV1.3 plasmid (Schwenzer et al, 2024), human CaV1.3 cDNA (accession number NM_001128840.2) was de-novo synthesized and assembled into a HaloTag vector (Promega G7721) using restriction cloning, thus encoding an N-terminal fusion of CaV1.3 to HaloTag linked by a ‘GGS’ sequence. The analogous GFP-CaV1.3 plasmid was generated by exchanging the HaloTag sequence for mEGFP using restriction cloning. The RBP2 construct with an mKATE2 tag comprised of an insert coding for mouse RBP2 (accession ID NP_001074857.1) with a p2A cleavage site followed by a C-terminal mKATE2 cloned inframe into a f(syn)w-mKATE2-p2A vector driven by a CMV promoter-enhancer and containing downstream WPRE sequence. The untagged version was made by cloning the RBP2 cDNA from this plasmid and replacing RIBEYE-GFP in the pAAV-CMV-HBA-RIBEYE-GFP-WPRE plasmid by in-fusion cloning (Takara Bio USA, Inc.).

Immunocytochemistry and Imaging

HEK293 cells plated on poly-L-lysine coated coverslips were fixed with 99% chilled methanol at −20°C for 2 minutes as previously described (Picher et al, 2017b). The coverslips were washed thoroughly three times with PBS at room temperature (5 – 10 min). Blocking and permeabilization was performed with GSDB (goat serum dilution buffer: 16% normal goat serum, 450 mM NaCl, 0.3% Triton X-100, 20mM phosphate buffer, pH ∼7.4) for 45-60 minutes at room temperature. Samples were incubated with respective primary antibodies (diluted in GSDB, refer to Table 2) overnight at 4°C or for 2 hours at RT. The samples were then washed three times (5–10 minutes each) with wash buffer (450 mM NaCl, 0.3% Triton X-100, 20mM phosphate buffer, pH ∼ 7.4). Incubation with appropriate secondary antibodies (also diluted in GSDB refer to Table 2) was performed for 1 hour at room temperature in a light-protected wet chamber. Lastly, the coverslips were washed three times with wash buffer (5-10 minutes each) and one final time with PBS, before mounting onto glass slides with a drop of fluorescence mounting medium (Mowiol 4-88, Carl Roth). For live cell imaging of cells transfected with Halo-CaV1.3 (Fig. 4B, C), cells plated on glass-bottomed dishes were treated with Janelia Fluor 646 HaloTag Ligand (Promega, Cat. No. GA1120) at a final labelling concentration of 200nM (in cell culture media). Cells were incubated with the ligand for ∼60 min at 37°C with 5% (v/v) CO2 saturation, after which the media was replaced with an equal volume of fresh warm culture media.

Images from fixed and live samples were acquired in confocal/ STED mode using an Olympus IX83 inverted microscope combined with an Abberior Instruments Expert Line STED microscope (Abberior Instruments GmbH). We used lasers at 488, 561, and 633 nm for excitation and a 775 nm (1.2W) laser for stimulated emission depletion. 1.4 NA 100X or 20X oil immersion objectives were used for fixed samples, and a water immersion 60X objective was used for live samples. Confocal stacks or single sections were acquired using Imspector Software (pixel size = 80 X 80 nm along xy, 200 nm along z). For 2D-STED images, a pixel size of 15 X 15 nm (in xy) was used. Images in Supplementary Fig. 2A were acquired using a Leica SP8 confocal microscope (Leica Microsystems, Germany). All acquired images were visualized using NIH ImageJ software and adjusted for brightness and contrast. Samples and their corresponding controls were processed in parallel using identical staining protocols, laser excitation powers and microscope settings.

Ultrastructural analysis using cryo-correlative electron-light microscopy

HEK293 cells were transfected with RIBEYE-GFP and untagged palm-Bassoon constructs as described above. The next day, the cells were detached using Accutase (Sigma-Aldrich, A6964) and replated on electron microscopy grids (R2/2, Au 200 mesh, 100 holey SiO2 film; Quantifoil) at a density of 20,000 cells/grid. After 24 hours, grids were vitrified by plunge-freezing in liquid ethane/propane using a house-made manual plunger. Cryo-focused ion beam milling and cryo-correlative light and electron microscopy were performed as previously described (Rigort et al, 2010; Pierson et al, 2024). In brief, upon transferring into an Aquilos 2 focused-ion beam/scanning electron microscope (ThermoFisher Scientific), the samples were imaged with an integrated fluorescent light microscope (ThermoFischer Scientific) at 470 nm excitation wavelength to locate cells expressing peripheral GFP, which corresponded to membrane-localised SyRibbons. Milling was performed at a 9° angle at these areas, using ion currents between 500 nA and 100 nA. During milling, the GFP fluorescence was monitored additionally when the lamellae reached 2 µm and 150 nm thickness to ensure that the structures of interest were captured. All fluorescent data is shown as maximum intensity projections of 4 µm z-stacks. For visualization purposes, the fluorescent image shown in Fig. 3A was corrected using the rolling ball background subtraction algorithm implemented in Fiji (Schindelin et al, 2012; Sternberg, 1983). After milling, lamellae were transferred into a Krios G4 cryo-transmission electron microscope (cryo-TEM; ThermoFisher Scientific). Correlation of fluorescence on 150 nm lamellae and 8700x TEM overview revealed electron-dense structures at GFP puncta. Tilt series of these densities were collected at 33k magnification using a dose-symmetric scheme (from −45° to +63°, 9° pretilt, 3° increments) and with a 20 eV energy filter. The total electron dose was 120 e-/Å2. The data was acquired using SerialEM (Mastronarde, 2005) running PACEtomo script (Eisenstein et al, 2023). Data pre-processing and tomogram reconstruction from tilt series were done automatically using an in-house script implementing MotionCor2, CTFFIND4 and IMOD (https://github.com/rubenlab/snartomo; Zheng et al, 2017; Rohou & Grigorieff, 2015; Mastronarde, 1997). To enhance contrast, a Wiener-like filter was applied (https://github.com/dtegunov/tom_deconv) on the tomograms.

Patch clamp and calcium imaging

CaV1.3 stably expressing HEK293 cells were plated on coverslips till about 40-50% confluency was achieved, following which they were induced for CaV1.3α1-D expression by incubating them for 24-48 hours in induction media supplemented with tetracycline. For transfected cells, transfection was performed after 24hrs of induction in the induction media as explained above. Cells were recorded in extracellular solution containing (in mM): 150 choline-Cl, 10 HEPES, 1 MgCl2 and 10 CaCl2; pH was adjusted to 7.3 with CsOH and osmolality was 300-310 mOsm/kg. For electrophysiological recordings in Supplementary Fig. 5 and 6, pipette solution contained (in mM): 140 NMDG, 10 NaCl, 10 HEPES, 5 EGTA, 3 Mg-ATP and 1 MgCl2. pH was adjusted to 7.3 with methanesulfonic acid and osmolality was around 290 mOsm/kg. For Ca2+ imaging in Fig. 7, the pipette solution contained (in mM): 111 Cs-glutamate, 1 MgCl2, 1 CaCl2, 10 EGTA, 13 TEA-Cl, 20 HEPES, 4 Mg-ATP, 0.3 Na-GTP, 1 L-glutathione, 0.1 Calbryte590 (AAT Bioquest, Cat. No. 20706); pH was adjusted to 7.3, osmolality to 290 mOsm/kg. Resistance of patch pipettes was 3–7 MΩ. HEK293 cells were rupture patch clamped with EPC-10 amplifier (HEKA electronics, Germany) controlled by Patchmaster software at room temperature, as described previously (Picher et al, 2017b, 2017a). Cells were kept at a holding potential of −91.2 or −88 mV. All voltages were corrected for liquid junction potential offline (21.2 mV or 18 mV). Currents were leak corrected using a p/10 protocol. Recordings were discarded when leak current exceeded −50 pA, Rs exceeded 15 MΩ or offset potential fluctuated more than 5 mV. Current voltage (IV) relations were recorded around 1 minute after rupturing the cell, by applying increasing step-depolarisation pulses of voltage ranging from −86.2 mV to 58.8 mV, in steps of 5 mV. Depolarisation pulses of 500 ms were used for measure CaV1.3 inactivation in Supplementary Fig. 6.

For Ca2+ imaging experiments, we used a Yokogawa CSU-X1A spinning-disk confocal scanner, mounted on an upright microscope (Zeiss Examiner) with a 63X, 1.0 NA objective (W Plan-Apochromat, Zeiss). Images were acquired using an Andor Zyla sCMOS 4.2 camera, controlled by VisiView 5.0 software (Visitron Systems GmbH). GFP, mKATE2 and the low-affinity Ca2+ indicator Calbryte590 were excited by a VS Laser Merge System (Visitron Systems) with 488 and 561 nm laser lines. Spinning disk was set to 5000 rpm. mKATE2 positive cells showing peripheral GFP expression (indicative of RIBEYE and palm-Bassoon co-expression) were identified and used for recordings. The cell was loaded for approximately a minute with Calbryte590 dye, after which we applied step depolarizations of 20 ms from −83 to +62 mV in 5-mV step increments to measure Ca2+ current (density)–voltage relations. Next, a central plane of the cell was selected and the GFP fluorescence (green channel) of the cell showing SyRibbons was imaged. This was immediately followed by imaging of the increments of Calbryte590 fluorescence (red channel) in the same plane triggered by a depolarizing pulse to +2 mV for 500 ms (frame rate = 20 Hz). For Ca2+ imaging experiments, we also included occasionally occurring HEK293 cells which were electrically coupled with other neighbouring HEK293 cells. For such cells, we could not fully compensate for the slow capacitive currents and we did not use them for electrophysiological analysis.

Data analysis

Analysis of confocal images

All images were processed using NIH ImageJ software to make z-projections and multi-channel composites, and figures were created using Adobe Illustrator. For analysis of membrane distribution, a region of interest (ROI) of 1 µm thickness was drawn along the periphery of the cell which was identified either using Na, K ATPase α1 immunofluorescence or by enhancing the contrast in images of transfected cells so that the cell boundary becomes distinguishable. Mean pixel intensity was measured along this ROI and for the area enclosed within the ROI. The ratio of the mean pixel intensity (periphery: inside) per cell was reported as an average from at least three central planes; we did not use the basal and top plane. For volumetric fits of SyRibbons, IHC ribbons and Ca2+ channel clusters, we used the inbuilt surface detection algorithm from Imaris 9.6 (Oxford Instruments). The surface detail was set to 0.16 µm and largest sphere diameters for background subtraction were set as follows: for RIBEYE/CtBP2 0.3 µm, for Bassoon 0.1 µm and for CaV1.3 0.08 µm. Thresholding was performed based on the quality of immunofluorescence to ensure all discernible surfaces were detected. All surfaces with less than 10 voxels were filtered out. A manual check was done to include undetected surfaces, remove surfaces not localizing within cell of interest, and split surfaces that would occasionally be clubbed together. For analysis of SyRibbons, only RIBEYE surfaces colocalizing with palm-Bassoon surfaces were used for quantifications (at 0.06 µm or less). Occasionally we would detect surfaces with volumes larger than 2 µm3 (usually cytosolic) or smaller than 0.02 µm3, which were not considered for analysis. For CaV1.3 cluster volumetric analysis in tetra-transfected cells, clusters classified as +SyRibbons were at 0.045 µm or less from palm-Bassoon positive RIBEYE surfaces. All other clusters were classified as –SyRibbons. Only clusters localized at the periphery of the cell were used for analysis. Pixel-based intensity correlation analysis was performed using the Coloc 2 plugin on ImageJ (http://imagej.net/Coloc_2). Analysis was performed on 1 µm thick regions of interests marking the periphery of the cell in at least three central planes (excluding the basal and top plane). Costes method was implemented with 100 repetitions to test for statistical significance of colocalisation coefficients (Costes et al, 2004).

Analysis of patch clamp and Ca2+ imaging data

Analysis of electrophysiology data and figure preparation was performed using Igor Pro 6 and 7 (WaveMetrics Inc.) and final figures were compiled using Adobe Illustrator. For analyzing current density-voltage relations (IVs), the evoked Ca2+ current was averaged from 5 to 10 ms after depolarization start. Cells expressing current amplitudes less than −49 pA and with current density greater than −60 pA/pF were not taken into account for analysis.

Ca2+ imaging data analysis was performed using a customized script in ImageJ. Briefly, we performed background subtraction in time series with Calbryte590 signal and then averaged 5 frames before stimulation and subtracted these from an average of three frames during stimulation to obtain ΔF images. For visualization, the images were smoothened (3 X 3 unweighted smoothening) and a composite was created by overlaying the ΔF images and the RIBEYE-GFP channel. For analyzing ΔF/F0, we drew circular regions of interest (ROIs) of 2µm diameter at sites with and without SyRibbons in the background subtracted time series. The ΔFmax/F0 was calculated as the of average ΔF/F0 in first three frames of peak Ca2+ signal intensity at stimulation onset in each ROI. To minimize bias while assigning regions as with or without SyRibbons, ROIs were assigned in the RIBEYE-GFP channel, without looking at the corresponding Calbryte590 signal.

Statistics

Data sorting and statistics were performed using MS Excel, Igor Pro 7 and/or GraphPad Prism 10. Numerical data is represented mainly as mean ± standard deviation (SD). The normality of data was assessed using Jarque-Bera test and Kolmogorov-Smirnov test, and the equality of variances was checked using F-test. When comparing two samples, a two-tailed unpaired Student’s t-test (or paired t-test for Fig. 7J) was performed for normally distributed samples with equal variances. If the conditions for normality and equality of variances were not met, an unpaired Mann-Whitney-Wilcoxon test was performed. When comparing multiple samples, a one-way ANOVA with post hoc Tukey’s test was used for normally distributed data with equal variances. For samples which were not normally distributed, we used Kruskal-Wallis test with post hoc Dunn’s multi-comparison correction. All box plots are depicted with crosses representing mean values, central bands indicating median, whiskers for 90/10 percentiles, boxes for 75/25 percentiles and individual data points overlaid. Non-significant differences have been indicated as n.s., while significant P values have been depicted as *P for < 0.05, **P < 0.01, ***P < 0.001, and ****P < 0.0001.